---
title: "NL"
author: "Lasse Steffensen"
date: "2026-03-02"
output: html_document
---

Pre
```{r}
setwd("C:/Users/lasse/Desktop/wd/25/01_05_cor/NL")


library(tidyverse)
library(scales)
library(ggVennDiagram)




#
meso_mags <- read.csv("C:/Users/lasse/Desktop/wd/25/coverm_rel_abun/meso_05_mags.tsv", header = TRUE, sep = "\t") %>%
  select(-rel_abun, -mean, -cov_frac)

thermo_mags <- read.csv("C:/Users/lasse/Desktop/wd/25/coverm_rel_abun/thermo_05_mags.tsv", header = TRUE, sep = "\t") %>%
  select(-rel_abun, -mean, -cov_frac)

MiDAS_GTDB_tax <- read.csv("C:/Users/lasse/Desktop/wd/25/01_05_cor/write_files/GTDB_MiDAS_tax.tsv", header = TRUE, sep = "\t")

module_completeness_all <- read.csv("C:/Users/lasse/Desktop/wd/25/df/26/module_completeness_all_dereps.tsv", header = TRUE, sep = "\t")


# Netherlands sample data
metadata <- read.csv("C:/Users/lasse/Desktop/wd/25/df/26/NL_data/metadata_1.txt", header = TRUE, sep = "\t")

otu_norm <- read.csv("C:/Users/lasse/Desktop/wd/25/df/26/NL_data/otutable_normalised.tsv", header = TRUE, sep = "\t") %>%
  select(OTU, barcode25, Kingdom, Phylum, Class, Order, Family, Genus, Species) %>%
  rename(rel_abun = barcode25) %>%
  mutate(Kingdom = str_trim(Kingdom)) %>%
  mutate(Phylum = str_trim(Phylum)) %>%
  mutate(Class = str_trim(Class)) %>%
  mutate(Order = str_trim(Order)) %>%
  mutate(Family = str_trim(Family)) %>%
  mutate(Genus = str_trim(Genus)) %>%
  mutate(Species = str_trim(Species))

midas_barcode_map <- read.csv("C:/Users/lasse/Desktop/wd/25/df/26/NL_data/usearch_midas_derep_out.tsv", header = FALSE, sep = "\t") %>%
  select(-V2, -V3) %>%
  rename(
    barcode = V1,
    midas_tax = V4
  )


# Sum rel_abun per distinct species in otu_norm
otu_norm_distinct <- otu_norm %>%
  group_by(Species) %>%
  mutate(rel_abun = sum(rel_abun, na.rm = TRUE)) %>%
  ungroup() %>%
  distinct(Species, .keep_all = TRUE) %>%
  filter(rel_abun > 0.05)


# Parse tax in midas_barcode_map
midas_barcode_map <- midas_barcode_map %>%
  separate_rows(midas_tax, sep = ",") %>%
  extract(midas_tax, into = c("rank", "name"), regex = "^([a-z]):(.*)$", remove = TRUE) %>%
  mutate(
    rank = recode(rank, d = "k"),
    value = paste0(rank, "__", name)
  ) %>%
  select(-name) %>%
  pivot_wider(
    names_from  = rank,
    values_from = value
  ) %>%
  rename(
    Kingdom = k,
    Phylum  = p,
    Class   = c,
    Order   = o,
    Family  = f,
    Genus   = g,
    Species = s
  ) %>%
  select(-"NA")


# Check how many species are not actually in midas_barcode_map
db_species  <- midas_barcode_map %>% distinct(Species)
otu_species <- otu_norm_distinct %>% distinct(Species)

# how many OTU species are present in database
n_present <- otu_species %>%
  semi_join(db_species, by = "Species") %>%
  nrow()

# how many are missing from database
n_missing <- otu_species %>%
  anti_join(db_species, by = "Species") %>%
  nrow()

n_present
n_missing




```

Barplot: Species-level contribution to relative abundance
```{r}

library(dplyr)

# 1) Vector of species present in BOTH dataframes
shared_species <- intersect(db_species$Species,
                            otu_species$Species)

# 2) Total relative abundance
total_rel_abun <- sum(otu_norm_distinct$rel_abun, na.rm = TRUE)

# 3) Relative abundance explained by shared species
explained_rel_abun <- otu_norm_distinct %>%
  filter(Species %in% shared_species) %>%
  summarise(total = sum(rel_abun, na.rm = TRUE)) %>%
  pull(total)

# 4) Proportion explained
proportion_explained <- explained_rel_abun / total_rel_abun

proportion_explained

```


Most abundant species (shared one's indicated)
```{r}


# top 30 species by rel_abun (from otu_norm_distinct)
top30 <- otu_norm_distinct %>%
  arrange(desc(rel_abun)) %>%
  slice_head(n = 30) %>%
  pull(Species)

# build stacked-bar data: each top30 species kept, all others collapsed to "Other"
plot_df <- otu_norm_distinct %>%
  mutate(group = if_else(Species %in% top30, Species, "Other")) %>%
  group_by(group) %>%
  summarise(rel_abun = sum(rel_abun, na.rm = TRUE), .groups = "drop") %>%
  mutate(group = factor(group, levels = c(top30, "Other")))

# colours (top30 coloured; Other grey)
cols <- c(setNames(hue_pal()(length(top30)), top30),
          Other = "grey70")

# labels with asterisk for shared species (legend only)
label_map <- setNames(
  ifelse(names(cols) %in% shared_species, paste0(names(cols), " *"), names(cols)),
  names(cols)
)

ggplot(plot_df, aes(x = "", y = rel_abun, fill = group)) +
  geom_col(width = 0.6) +
  scale_fill_manual(values = cols, breaks = names(cols), labels = label_map) +
  theme_bw() +
  labs(x = NULL, y = "Relative abundance", fill = "Species (* = shared)") +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())

```


Top 30 genera
```{r}

library(dplyr)
library(ggplot2)
library(scales)

# 1) Sum rel_abun per genus
otu_genus <- otu_norm %>%
  group_by(Genus) %>%
  summarise(rel_abun = sum(rel_abun, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(rel_abun))

# 2) Top 30 genera
top30_g <- otu_genus %>%
  slice_head(n = 30) %>%
  pull(Genus)

# 3) Collapse remainder to "Other" for a single stacked bar
plot_df <- otu_genus %>%
  mutate(group = if_else(Genus %in% top30_g, Genus, "Other")) %>%
  group_by(group) %>%
  summarise(rel_abun = sum(rel_abun, na.rm = TRUE), .groups = "drop") %>%
  mutate(group = factor(group, levels = c(top30_g, "Other")))

# 4) Colours + plot
cols <- c(setNames(hue_pal()(length(top30_g)), top30_g),
          Other = "grey70")

ggplot(plot_df, aes(x = "", y = rel_abun, fill = group)) +
  geom_col(width = 0.6) +
  scale_fill_manual(values = cols, breaks = names(cols)) +
  theme_bw() +
  labs(x = NULL, y = "Relative abundance", fill = "Genus") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

```


MiDAS taxonomy overlap: Netherlands >< my samples
```{r}

# Taxonomy dataframes for meso and thermo
vm <- c("NNF_AD05_mmlongv091.bin.1.449", "NNF_AD17_mmlongv091.bin.1.120", "NNF_AD07_mmlongv091.bin.1.54", "NNF_AD03_mmlongv091.bin.1.182")

meso_mags_tax <- meso_mags %>%
  left_join(MiDAS_GTDB_tax, by = c("Genome" = "FLASV")) %>%
  filter(!Genome %in% vm)

vt <- c("NNF_AD03_mmlongv091.bin.1.182", "BC14.bin.2.701", "NNF_AD07_mmlongv091.bin.1.54", "BC13.bin.3.839", "NNF_AD14_mmlongv091.bin.1.148", "NNF_AD17_mmlongv091.bin.1.120", "NNF_AD09_mmlongv091.bin.1.346")

thermo_mags_tax <- thermo_mags %>%
  left_join(MiDAS_GTDB_tax, by = c("Genome" = "FLASV")) %>%
  filter(!Genome %in% vt)

# Reformat meso and thermo to match otu_norm tax format
meso_mags_tax_s <- meso_mags_tax %>%
  mutate(s = paste0("s__", s))

thermo_mags_tax_s <- thermo_mags_tax %>%
  mutate(s = paste0("s__", s))


# Create Venn diagram
# species sets (unique, non-missing)
sets <- list(
  `NL sample` = otu_norm_distinct %>%
    transmute(species = Species) %>%
    filter(!is.na(species), species != "") %>%
    distinct() %>%
    pull(species),

  `Meso` = meso_mags_tax_s %>%
    transmute(species = s) %>%
    filter(!is.na(species), species != "") %>%
    distinct() %>%
    pull(species),

  `Thermo` = thermo_mags_tax_s %>%
    transmute(species = s) %>%
    filter(!is.na(species), species != "") %>%
    distinct() %>%
    pull(species)
)

ggVennDiagram(sets, label_alpha = 0) +
  theme_light() +
  labs(title = "Species-level overlap")


```

Genus-level overlap (Venn diagram)
```{r}

# Reformat meso and thermo to match otu_norm tax format
meso_mags_tax_g <- meso_mags_tax %>%
  mutate(g = paste0("g__", g))

thermo_mags_tax_g <- thermo_mags_tax %>%
  mutate(g = paste0("g__", g))


# Create Venn diagram
# Genus sets (unique, non-missing)
sets <- list(
  `NL sample` = otu_norm_distinct %>%
    transmute(genus = Genus) %>%
    filter(!is.na(genus), genus != "") %>%
    distinct() %>%
    pull(genus),

  `Meso` = meso_mags_tax_g %>%
    transmute(genus = g) %>%
    filter(!is.na(genus), genus != "") %>%
    distinct() %>%
    pull(genus),

  `Thermo` = thermo_mags_tax_g %>%
    transmute(genus = g) %>%
    filter(!is.na(genus), genus != "") %>%
    distinct() %>%
    pull(genus)
)

ggVennDiagram(sets, label_alpha = 0) +
  theme_light() +
  labs(title = "Genus-level overlap")

```


```{r}


```


```{r}


```


```{r}


```


```{r}


```


```{r}


```


```{r}


```


```{r}


```

