---
title: "SBP_deg"
author: "Lasse Steffensen"
date: "2026-02-17"
output: html_document
---

--------------------------------------------------------------------------------
                                      Pre
--------------------------------------------------------------------------------

```{r}

setwd("C:/Users/lasse/Desktop/wd/25/01_05/SBP_deg")

library(tidyverse)
library(readxl)
library(patchwork)
library(scales)
library(grid)
library(viridisLite)
library(scico)



file <- "C:/Users/lasse/Desktop/wd/25/df/metabolism_summary_26.xlsx"
metabolism_summary <- read_excel(file, sheet = 2) %>%
  mutate(gene_id = as.character(gene_id))


meso_mags_tax <- read.csv(
  "C:/Users/lasse/Desktop/wd/25/01_05/write_files/05/meso_mags_tax.tsv",
  header = TRUE, sep = "\t"
)

thermo_mags_tax <- read.csv(
  "C:/Users/lasse/Desktop/wd/25/01_05/write_files/05/thermo_mags_tax.tsv",
  header = TRUE, sep = "\t"
)

# original MAG IDs
mags_meso_old   <- as.character(meso_mags_tax$Genome)
mags_thermo_old <- as.character(thermo_mags_tax$Genome)


# Import name map
name_map <- read.csv("C:/Users/lasse/Desktop/wd/25/01_05/module_completeness/name_map.tsv", header = TRUE, sep = "\t")




# Create named rename vector
rename_vec <- setNames(name_map$new, name_map$old)


# Extract gene counts
gene_counts_all_old <- metabolism_summary %>%
  mutate(gene_id = as.character(gene_id)) %>%
  select(gene_id, 6:ncol(.)) %>%
  mutate(across(-gene_id, ~ parse_number(as.character(.x))))


# Subset into samples
gene_counts_meso <- gene_counts_all_old %>%
  select(gene_id, any_of(mags_meso_old))

gene_counts_thermo <- gene_counts_all_old %>%
  select(gene_id, any_of(mags_thermo_old))


# Rename columns into tax labels
cols_meso <- intersect(names(gene_counts_meso), names(rename_vec))

gene_counts_meso <- gene_counts_meso %>%
  rename_with(~ rename_vec[.x], all_of(cols_meso))

names(gene_counts_meso) <- make.unique(names(gene_counts_meso), sep = "_MAG")


cols_thermo <- intersect(names(gene_counts_thermo), names(rename_vec))

gene_counts_thermo <- gene_counts_thermo %>%
  rename_with(~ rename_vec[.x], all_of(cols_thermo))

names(gene_counts_thermo) <- make.unique(names(gene_counts_thermo), sep = "_MAG")


# Add rel abun
meso_mags_tax <- meso_mags_tax %>%
  left_join(name_map, by = c("Genome" = "old"))

thermo_mags_tax <- thermo_mags_tax %>%
  left_join(name_map, by = c("Genome" = "old"))



```

        --------------------------------------------------------------
    -----------------------------------------------------------------------
--------------------------------------------------------------------------------
                                     
                                      Pectin

--------------------------------------------------------------------------------
    -----------------------------------------------------------------------
        --------------------------------------------------------------
    

Make lists of target gene_ids
```{r}



# Make classifications

gene_classes <- tibble(
  gene_id = c(
    # Pectin backbone cleavage
    "GH28", "PL1", "PL10", "PL11", "PL2", "PL22", "PL26", "PL3", "PL4", "PL9",

    # Pectin oligo cleavage
    "CE12", "CE8", "GH105", "GH138", "GH143", "GH4", "GH139", "GH106", "GH78",
    
    # Pectic galactan backbone cleavage
    "GH53",

    # Pectic galactan oligo cleavage
    "GH147", "GH165", "GH35", "GH42", "GH59",
    
    # Arabinan oligo cleavage
    "GH51", "GH54", "GH62", "GH93"
  ),
  class = c(
    rep("Pectin backbone cleavage", 10),
    rep("Pectin oligo cleavage", 9),
    rep("Pectic galactan backbone cleavage", 1),
    rep("Pectic galactan oligo cleavage", 5),
    rep("Arabinan oligo cleavage", 4)
  )
)


```


Obtain target gene_ids for target MAGs
```{r}



# Filter for pectin CAZYs
gene_count_meso_pectin <- gene_counts_meso %>%
  filter(gene_id %in% gene_classes$gene_id)

gene_count_thermo_pectin <- gene_counts_thermo %>%
  filter(gene_id %in% gene_classes$gene_id)



```


--------------------------------------------------------------------------------
                                      Heatmaps
--------------------------------------------------------------------------------


```{r}


meso_long <- gene_count_meso_pectin %>%
  pivot_longer(
    cols = -gene_id,
    names_to = "genome",
    values_to = "count"
  )


thermo_long <- gene_count_thermo_pectin %>%
  pivot_longer(
    cols = -gene_id,
    names_to = "genome",
    values_to = "count"
  )


p_meso <- ggplot(meso_long, aes(x = genome, y = gene_id, fill = count)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Count") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10),
    axis.title = element_blank()
  ) +
  ggtitle("Pectin CAZy repertoire – Mesophilic MAGs")


p_thermo <- ggplot(thermo_long, aes(x = genome, y = gene_id, fill = count)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Count") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10),
    axis.title = element_blank()
  ) +
  ggtitle("Pectin CAZy repertoire – Thermophilic MAGs")

p <- p_meso / p_thermo

ggsave(filename = "pectin/gene_count_heatmap.png", plot = p,
       width = 20, height = 12, dpi = 300)

```


--------------------------------------------------------------------------------
                           Per-gene id MAG amount barplot
--------------------------------------------------------------------------------


Class-separated
```{r}


# ------------------------------------------------------------
# Helper: count MAGs with count > 0 per gene_id
# ------------------------------------------------------------
count_mags_per_gene <- function(df_counts) {
  df_counts %>%
    mutate(across(-gene_id, ~ as.numeric(.x))) %>%
    pivot_longer(-gene_id, names_to = "Genome", values_to = "count") %>%
    mutate(present = !is.na(count) & count > 0) %>%
    group_by(gene_id) %>%
    summarise(n_MAGs = sum(present), .groups = "drop")
}

# helper: make labels fit in skinny facets
# "GH53" -> "GH\n53" (works nicely for GH/PL/CE codes)
stack_gene_label <- function(x) {
  x <- as.character(x)
  gsub("^([A-Z]+)([0-9]+)$", "\\1\n\\2", x)
}

# ------------------------------------------------------------
# Plot function: facet by class + class colours from a gradient
# ------------------------------------------------------------
plot_cazy_by_class <- function(df_counts, gene_classes, title, y_max = NULL,
                               gradient_anchors = c("grey80", "steelblue4"),
                               label_stack = TRUE) {

  hits <- count_mags_per_gene(df_counts)

  gene_order <- gene_classes %>%
    mutate(gene_id = as.character(gene_id)) %>%
    pull(gene_id)

  plot_df <- gene_classes %>%
    mutate(
      gene_id = as.character(gene_id),
      class   = str_squish(class)
    ) %>%
    left_join(hits, by = "gene_id") %>%
    mutate(
      n_MAGs  = replace_na(n_MAGs, 0L),
      class   = fct_inorder(class),
      gene_id = factor(gene_id, levels = unique(gene_order))
    )

  class_levels <- gene_classes %>%
    mutate(class = str_squish(class)) %>%
    distinct(class) %>%
    pull(class)

  plot_df <- plot_df %>%
    mutate(class = factor(as.character(class), levels = class_levels))

  pal_fun <- colorRampPalette(gradient_anchors)
  class_cols <- setNames(pal_fun(length(class_levels)), class_levels)

  p <- ggplot(plot_df, aes(x = gene_id, y = n_MAGs, fill = class)) +
    geom_col(width = 0.9) +
    facet_grid(. ~ class, scales = "free_x", space = "free_x", switch = "x") +
    scale_fill_manual(values = class_cols, drop = FALSE) +
    geom_text(aes(label = n_MAGs), vjust = -0.3, size = 2.5) +
    labs(
      x = NULL,
      y = "Number of MAGs (gene count > 0)",
      title = title,
      fill = "Class"
    ) +
    theme_bw() +
    theme(
      axis.text.x = element_text(
        angle = 0,
        hjust = 0.5, vjust = 1,
        size = 7,
        lineheight = 0.9
      ),
      axis.ticks.x = element_line(linewidth = 0.25),
      panel.spacing.x = unit(0.9, "lines"),
      legend.position = "none",
      strip.placement = "outside",
      strip.background = element_rect(fill = "grey95", colour = NA),
      plot.margin = margin(t = 5.5, r = 5.5, b = 18, l = 5.5)
    ) +
    scale_y_continuous(
      expand = expansion(mult = c(0, 0.10))
    ) +
    coord_cartesian(ylim = c(0, 20), clip = "off")

  if (isTRUE(label_stack)) {
    p <- p + scale_x_discrete(labels = stack_gene_label)
  } else {
    p <- p + scale_x_discrete(drop = FALSE)
  }

  p
}

# ------------------------------------------------------------
# Run for meso (blue gradient) and thermo (red gradient)
# ------------------------------------------------------------
p_meso <- plot_cazy_by_class(
  df_counts = gene_count_meso_pectin,
  gene_classes = gene_classes,
  title = "Number of mesophilic MAGs encoding each pectin CAZy (count > 0)",
  y_max = 20,
  gradient_anchors = c("cornflowerblue", "royalblue2", "royalblue3", "royalblue4", "midnightblue"),
  label_stack = TRUE
)

p_thermo <- plot_cazy_by_class(
  df_counts = gene_count_thermo_pectin,
  gene_classes = gene_classes,
  title = "Number of thermophilic MAGs encoding each pectin CAZy (count > 0)",
  y_max = 20,
  gradient_anchors = c("sandybrown", "sienna1", "sienna2", "sienna3", "sienna", "sienna4"),
  label_stack = TRUE
)

p <- p_meso / p_thermo
p



ggsave(filename = "pectin/mag_gene_counts_barplot.png", plot = p,
       width = 20, height = 12, dpi = 300)


```


Attribution plot (top 10 by gene count + X)
```{r}

special_meso <- c("UBA3950 Nemesis")


make_stacked_contrib_plot <- function(df_counts,
                                      gene_classes = NULL,
                                      top_n = 10,
                                      title = "",
                                      mode = c("proportion", "count"),
                                      facet_by_class = TRUE,
                                      genome_palette = NULL,
                                      other_colour = "grey70",
                                      # abundance mapping:
                                      mags_tax = NULL,
                                      name_col = "new",
                                      rel_abun_col = "rel_abun",
                                      rel_abun_digits = 2,
                                      # NEW:
                                      highlight_genomes = character(0)) {

  mode <- match.arg(mode)

  # Wide -> long
  long <- df_counts %>%
    dplyr::mutate(dplyr::across(-gene_id, ~ as.numeric(.x))) %>%
    tidyr::pivot_longer(-gene_id, names_to = "Genome", values_to = "count") %>%
    dplyr::mutate(count = tidyr::replace_na(count, 0))

  # Top N by total
  top_genomes <- long %>%
    dplyr::group_by(Genome) %>%
    dplyr::summarise(total = sum(count), .groups = "drop") %>%
    dplyr::arrange(dplyr::desc(total)) %>%
    dplyr::slice_head(n = top_n) %>%
    dplyr::pull(Genome)

  # NEW: combine + preserve order (top first, then highlighted that aren't already in top)
  keep_genomes <- c(top_genomes, setdiff(highlight_genomes, top_genomes))

  # Collapse others (but keep highlighted)
  long2 <- long %>%
    dplyr::mutate(Genome_grp = dplyr::if_else(Genome %in% keep_genomes, Genome, "Other"))

  # Aggregate
  plot_df <- long2 %>%
    dplyr::group_by(gene_id, Genome_grp) %>%
    dplyr::summarise(count = sum(count), .groups = "drop")

  # Proportion mode
  if (mode == "proportion") {
    plot_df <- plot_df %>%
      dplyr::group_by(gene_id) %>%
      dplyr::mutate(
        denom = sum(count),
        prop  = dplyr::if_else(denom > 0, count / denom, 0)
      ) %>%
      dplyr::ungroup() %>%
      dplyr::select(-denom)
  }

  # Join class info
  if (!is.null(gene_classes)) {
    gene_order <- gene_classes %>% dplyr::pull(gene_id) %>% as.character()
    class_levels <- gene_classes %>%
      dplyr::mutate(class = stringr::str_squish(class)) %>%
      dplyr::distinct(class) %>%
      dplyr::pull(class)

    plot_df <- plot_df %>%
      dplyr::left_join(
        gene_classes %>%
          dplyr::mutate(gene_id = as.character(gene_id),
                        class = stringr::str_squish(class)),
        by = "gene_id"
      ) %>%
      dplyr::mutate(
        class = factor(class, levels = class_levels),
        gene_id = factor(as.character(gene_id),
                         levels = unique(gene_order))
      )
  } else {
    plot_df <- plot_df %>%
      dplyr::mutate(gene_id = factor(as.character(gene_id)))
  }

  # Factor levels for legend order
  genome_levels <- c(keep_genomes, "Other")
  plot_df <- plot_df %>%
    dplyr::mutate(Genome_grp = factor(Genome_grp, levels = genome_levels))

  # -------------------------
  # Colours
  # -------------------------
  n_keep <- length(keep_genomes)

  if (is.null(genome_palette)) {
    pal_fun <- grDevices::colorRampPalette(c("steelblue4", "goldenrod2", "firebrick3"))
    keep_cols <- stats::setNames(pal_fun(n_keep), keep_genomes)
  } else {
    if (length(genome_palette) < n_keep) {
      stop("Not enough colours supplied in genome_palette for top_n + highlight_genomes.")
    }
    keep_cols <- stats::setNames(genome_palette[seq_len(n_keep)], keep_genomes)
  }
  mag_cols <- c(keep_cols, Other = other_colour)

  # -------------------------
  # Legend labels with rel_abun for kept genomes
  # -------------------------
  legend_labels <- stats::setNames(genome_levels, genome_levels)

  if (!is.null(mags_tax)) {
    ab_map <- mags_tax %>%
      dplyr::select(
        Genome = dplyr::all_of(name_col),
        rel_abun = dplyr::all_of(rel_abun_col)
      ) %>%
      dplyr::filter(!is.na(Genome)) %>%
      dplyr::distinct(Genome, .keep_all = TRUE)

    ab_vec <- ab_map$rel_abun
    names(ab_vec) <- ab_map$Genome

    keep_labels <- vapply(
      keep_genomes,
      function(g) {
        ra <- ab_vec[[g]]
        if (is.null(ra) || is.na(ra)) {
          sprintf("%s (NA%%)", g)
        } else {
          sprintf(paste0("%s (%.", rel_abun_digits, "f%%)"), g, ra)
        }
      },
      character(1)
    )
    names(keep_labels) <- keep_genomes

    legend_labels <- c(keep_labels, Other = "Other")
  }

  # -------------------------
  # Plot
  # -------------------------
  yvar <- if (mode == "proportion") "prop" else "count"
  ylab <- if (mode == "proportion") "Proportion of total gene count"
  else "Total gene count (summed across selected MAGs)"

  p <- ggplot2::ggplot(plot_df, ggplot2::aes(x = gene_id, y = .data[[yvar]], fill = Genome_grp)) +
    ggplot2::geom_col(width = 0.9) +
    ggplot2::scale_fill_manual(
      values = mag_cols,
      breaks = genome_levels,
      labels = legend_labels,
      drop = FALSE
    ) +
    ggplot2::labs(
      x = "CAZy family (gene_id)",
      y = ylab,
      title = title,
      fill = "MAG / species"
    ) +
    ggplot2::theme_bw() +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, vjust = 1, size = 7),
      legend.position = "right"
    )

  if (facet_by_class && !is.null(gene_classes)) {
    p <- p +
      ggplot2::facet_grid(. ~ class, scales = "free_x", space = "free_x", switch = "x") +
      ggplot2::theme(
        panel.spacing.x = grid::unit(0.8, "lines"),
        strip.placement = "outside",
        strip.background = ggplot2::element_rect(fill = "grey95", colour = NA)
      )
  }

  p <- p +
    ggplot2::scale_y_continuous(
      labels = if (mode == "proportion") scales::percent_format(accuracy = 1) else waiver(),
      expand = ggplot2::expansion(mult = c(0, 0.05))
    ) +
    ggplot2::coord_cartesian(ylim = c(0, 45))

  return(p)
}



p_meso_stack <- make_stacked_contrib_plot(
  df_counts = gene_count_meso_pectin,
  gene_classes = gene_classes,
  top_n = 10,
  highlight_genomes = special_meso,   # <- THIS was missing
  mode = "count",
  mags_tax = meso_mags_tax,
  name_col = "new",
  rel_abun_col = "rel_abun",
  rel_abun_digits = 2,
  genome_palette = c(
    "orchid1", "purple", "slateblue", "slateblue4", "navy",
    "steelblue3", "mediumaquamarine",
    "seagreen1", "seagreen3", "olivedrab", "khaki4"
  ),
  other_colour = "grey80",
  title = "Mesophilic contribution to pectin CAZy gene counts"
)

p_thermo_stack <- make_stacked_contrib_plot(
  df_counts = gene_count_thermo_pectin,
  gene_classes = gene_classes,
  top_n = 10,
  mode = "count",
  mags_tax = thermo_mags_tax,    # NEW
  name_col = "new",
  rel_abun_col = "rel_abun",
  rel_abun_digits = 2,
  genome_palette = c(
    "black", "firebrick4", "firebrick3",
    "firebrick1", "gold3", "gold1", "darkorange1",
    "chocolate", "darkorange4", "darkred"
  ),
  other_colour = "grey80",
  title = "Thermophilic contribution to pectin CAZy gene counts"
)

p <- p_meso_stack / p_thermo_stack


ggsave(filename = "pectin/attribution_barplot.png", plot = p,
       width = 20, height = 12, dpi = 300)

```


Attribution plot (top 10 by abundance)
```{r}

make_stacked_contrib_plot <- function(df_counts,
                                      gene_classes = NULL,
                                      top_n = 10,
                                      title = "",
                                      mode = c("proportion", "count"),
                                      facet_by_class = TRUE,
                                      genome_palette = NULL,
                                      other_colour = "grey70",
                                      # abundance mapping:
                                      mags_tax = NULL,
                                      name_col = "new",
                                      rel_abun_col = "rel_abun",
                                      rel_abun_digits = 2) {

  mode <- match.arg(mode)

  # Wide -> long
  long <- df_counts %>%
    dplyr::mutate(dplyr::across(-gene_id, ~ as.numeric(.x))) %>%
    tidyr::pivot_longer(-gene_id, names_to = "Genome", values_to = "count") %>%
    dplyr::mutate(count = tidyr::replace_na(count, 0))

  # -------------------------
  # Select TOP_N GENOMES
  #   - Prefer rel_abun from mags_tax (most abundant)
  #   - Fall back to total gene counts if mags_tax is NULL
  # -------------------------
  if (!is.null(mags_tax)) {

    ab_map <- mags_tax %>%
      dplyr::select(
        Genome = dplyr::all_of(name_col),
        rel_abun = dplyr::all_of(rel_abun_col)
      ) %>%
      dplyr::filter(!is.na(Genome)) %>%
      dplyr::mutate(rel_abun = as.numeric(rel_abun)) %>%
      dplyr::distinct(Genome, .keep_all = TRUE)

    # Only keep genomes that exist in df_counts
    present_genomes <- unique(long$Genome)

    top_genomes <- ab_map %>%
      dplyr::filter(Genome %in% present_genomes) %>%
      dplyr::arrange(dplyr::desc(rel_abun)) %>%
      dplyr::slice_head(n = top_n) %>%
      dplyr::pull(Genome)

  } else {

    # fallback: top N by total gene contribution
    top_genomes <- long %>%
      dplyr::group_by(Genome) %>%
      dplyr::summarise(total = sum(count), .groups = "drop") %>%
      dplyr::arrange(dplyr::desc(total)) %>%
      dplyr::slice_head(n = top_n) %>%
      dplyr::pull(Genome)
  }

  keep_genomes <- top_genomes

  # Collapse others
  long2 <- long %>%
    dplyr::mutate(Genome_grp = dplyr::if_else(Genome %in% keep_genomes, Genome, "Other"))

  # Aggregate counts per gene_id and Genome_grp
  plot_df <- long2 %>%
    dplyr::group_by(gene_id, Genome_grp) %>%
    dplyr::summarise(count = sum(count), .groups = "drop")

  # Proportion mode
  if (mode == "proportion") {
    plot_df <- plot_df %>%
      dplyr::group_by(gene_id) %>%
      dplyr::mutate(
        denom = sum(count),
        prop  = dplyr::if_else(denom > 0, count / denom, 0)
      ) %>%
      dplyr::ungroup() %>%
      dplyr::select(-denom)
  }

  # Join class info
  if (!is.null(gene_classes)) {
    gene_order <- gene_classes %>% dplyr::pull(gene_id) %>% as.character()
    class_levels <- gene_classes %>%
      dplyr::mutate(class = stringr::str_squish(class)) %>%
      dplyr::distinct(class) %>%
      dplyr::pull(class)

    plot_df <- plot_df %>%
      dplyr::left_join(
        gene_classes %>%
          dplyr::mutate(gene_id = as.character(gene_id),
                        class = stringr::str_squish(class)),
        by = "gene_id"
      ) %>%
      dplyr::mutate(
        class = factor(class, levels = class_levels),
        gene_id = factor(as.character(gene_id),
                         levels = unique(gene_order))
      )
  } else {
    plot_df <- plot_df %>%
      dplyr::mutate(gene_id = factor(as.character(gene_id)))
  }

  # Factor levels for legend order
  genome_levels <- c(keep_genomes, "Other")
  plot_df <- plot_df %>%
    dplyr::mutate(Genome_grp = factor(Genome_grp, levels = genome_levels))

  # -------------------------
  # Colours
  # -------------------------
  n_keep <- length(keep_genomes)

  if (is.null(genome_palette)) {
    pal_fun <- grDevices::colorRampPalette(c("steelblue4", "goldenrod2", "firebrick3"))
    keep_cols <- stats::setNames(pal_fun(n_keep), keep_genomes)
  } else {
    if (length(genome_palette) < n_keep) {
      stop("Not enough colours supplied in genome_palette for top_n genomes.")
    }
    keep_cols <- stats::setNames(genome_palette[seq_len(n_keep)], keep_genomes)
  }

  mag_cols <- c(keep_cols, Other = other_colour)

  # -------------------------
  # Legend labels with rel_abun for kept genomes
  # -------------------------
  legend_labels <- stats::setNames(genome_levels, genome_levels)

  if (!is.null(mags_tax)) {
    ab_map2 <- mags_tax %>%
      dplyr::select(
        Genome = dplyr::all_of(name_col),
        rel_abun = dplyr::all_of(rel_abun_col)
      ) %>%
      dplyr::filter(!is.na(Genome)) %>%
      dplyr::mutate(rel_abun = as.numeric(rel_abun)) %>%
      dplyr::distinct(Genome, .keep_all = TRUE)

    ab_vec <- ab_map2$rel_abun
    names(ab_vec) <- ab_map2$Genome

    keep_labels <- vapply(
      keep_genomes,
      function(g) {
        ra <- ab_vec[[g]]
        if (is.null(ra) || is.na(ra)) {
          sprintf("%s (NA%%)", g)
        } else {
          sprintf(paste0("%s (%.", rel_abun_digits, "f%%)"), g, ra)
        }
      },
      character(1)
    )
    names(keep_labels) <- keep_genomes

    legend_labels <- c(keep_labels, Other = "Other")
  }

  # -------------------------
  # Plot
  # -------------------------
  yvar <- if (mode == "proportion") "prop" else "count"
  ylab <- if (mode == "proportion") "Proportion of total gene count"
  else "Total gene count (summed across selected MAGs)"

  p <- ggplot2::ggplot(plot_df, ggplot2::aes(x = gene_id, y = .data[[yvar]], fill = Genome_grp)) +
    ggplot2::geom_col(width = 0.9) +
    ggplot2::scale_fill_manual(
      values = mag_cols,
      breaks = genome_levels,
      labels = legend_labels,
      drop = FALSE
    ) +
    ggplot2::labs(
      x = "CAZy family (gene_id)",
      y = ylab,
      title = title,
      fill = "MAG / species"
    ) +
    ggplot2::theme_bw() +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, vjust = 1, size = 7),
      legend.position = "right"
    )

  if (facet_by_class && !is.null(gene_classes)) {
    p <- p +
      ggplot2::facet_grid(. ~ class, scales = "free_x", space = "free_x", switch = "x") +
      ggplot2::theme(
        panel.spacing.x = grid::unit(0.8, "lines"),
        strip.placement = "outside",
        strip.background = ggplot2::element_rect(fill = "grey95", colour = NA)
      )
  }

  p <- p +
    ggplot2::scale_y_continuous(
      labels = if (mode == "proportion") scales::percent_format(accuracy = 1) else waiver(),
      expand = ggplot2::expansion(mult = c(0, 0.05))
    ) +
    ggplot2::coord_cartesian(ylim = c(0, 45))

  return(p)
}


p_meso_stack <- make_stacked_contrib_plot(
  df_counts = gene_count_meso_pectin,
  gene_classes = gene_classes,
  top_n = 10,
  mode = "count",
  mags_tax = meso_mags_tax,      # now drives top_n selection by rel_abun
  name_col = "new",
  rel_abun_col = "rel_abun",
  rel_abun_digits = 2,
  genome_palette = c(
    "orchid1", "purple", "slateblue", "slateblue4", "navy",
    "steelblue3", "mediumaquamarine",
    "seagreen1", "seagreen3", "olivedrab"
  ),
  other_colour = "grey80",
  title = "Mesophilic contribution to pectin CAZy gene counts"
)

p_thermo_stack <- make_stacked_contrib_plot(
  df_counts = gene_count_thermo_pectin,
  gene_classes = gene_classes,
  top_n = 10,
  mode = "count",
  mags_tax = thermo_mags_tax,    # now drives top_n selection by rel_abun
  name_col = "new",
  rel_abun_col = "rel_abun",
  rel_abun_digits = 2,
  genome_palette = c(
    "black", "firebrick4", "firebrick3",
    "firebrick1", "gold3", "gold1", "darkorange1",
    "chocolate", "darkorange4", "darkred"
  ),
  other_colour = "grey80",
  title = "Thermophilic contribution to pectin CAZy gene counts"
)

p <- p_meso_stack / p_thermo_stack



ggsave(filename = "pectin/abun_attribution_barplot.png", plot = p,
       width = 20, height = 12, dpi = 300)

```


Abundance only Attribute plot
```{r}

make_stacked_contrib_plot <- function(df_counts,
                                      gene_classes = NULL,
                                      top_n = 10,
                                      title = "",
                                      mode = c("proportion", "count"),
                                      facet_by_class = TRUE,
                                      genome_palette = NULL,
                                      other_colour = "grey70",
                                      mags_tax = NULL,
                                      name_col = "new",
                                      rel_abun_col = "rel_abun",
                                      rel_abun_digits = 2) {

  mode <- match.arg(mode)

  long <- df_counts %>%
    dplyr::mutate(dplyr::across(-gene_id, ~ as.numeric(.x))) %>%
    tidyr::pivot_longer(-gene_id, names_to = "Genome", values_to = "count") %>%
    dplyr::mutate(count = tidyr::replace_na(count, 0))

  if (!is.null(mags_tax)) {

    ab_map <- mags_tax %>%
      dplyr::select(
        Genome = dplyr::all_of(name_col),
        rel_abun = dplyr::all_of(rel_abun_col)
      ) %>%
      dplyr::filter(!is.na(Genome)) %>%
      dplyr::mutate(rel_abun = as.numeric(rel_abun)) %>%
      dplyr::distinct(Genome, .keep_all = TRUE)

    present_genomes <- unique(long$Genome)

    genomes_with_any_gene <- long %>%
      dplyr::group_by(Genome) %>%
      dplyr::summarise(has_gene = any(count >= 1), .groups = "drop") %>%
      dplyr::filter(has_gene) %>%
      dplyr::pull(Genome)

    top_genomes <- ab_map %>%
      dplyr::filter(Genome %in% present_genomes, Genome %in% genomes_with_any_gene) %>%
      dplyr::arrange(dplyr::desc(rel_abun)) %>%
      dplyr::slice_head(n = top_n) %>%
      dplyr::pull(Genome)

  } else {

    top_genomes <- long %>%
      dplyr::group_by(Genome) %>%
      dplyr::summarise(total = sum(count), .groups = "drop") %>%
      dplyr::arrange(dplyr::desc(total)) %>%
      dplyr::slice_head(n = top_n) %>%
      dplyr::pull(Genome)
  }

  keep_genomes <- top_genomes

  long2 <- long %>%
    dplyr::mutate(Genome_grp = dplyr::if_else(Genome %in% keep_genomes, Genome, "Other"))

  plot_df <- long2 %>%
    dplyr::group_by(gene_id, Genome_grp) %>%
    dplyr::summarise(count = sum(count), .groups = "drop")

  if (mode == "proportion") {
    plot_df <- plot_df %>%
      dplyr::group_by(gene_id) %>%
      dplyr::mutate(
        denom = sum(count),
        prop  = dplyr::if_else(denom > 0, count / denom, 0)
      ) %>%
      dplyr::ungroup() %>%
      dplyr::select(-denom)
  }

  if (!is.null(gene_classes)) {
    gene_order <- gene_classes %>% dplyr::pull(gene_id) %>% as.character()
    class_levels <- gene_classes %>%
      dplyr::mutate(class = stringr::str_squish(class)) %>%
      dplyr::distinct(class) %>%
      dplyr::pull(class)

    plot_df <- plot_df %>%
      dplyr::left_join(
        gene_classes %>%
          dplyr::mutate(gene_id = as.character(gene_id),
                        class = stringr::str_squish(class)),
        by = "gene_id"
      ) %>%
      dplyr::mutate(
        class = factor(class, levels = class_levels),
        gene_id = factor(as.character(gene_id),
                         levels = unique(gene_order))
      )
  } else {
    plot_df <- plot_df %>%
      dplyr::mutate(gene_id = factor(as.character(gene_id)))
  }

  genome_levels <- c(keep_genomes, "Other")
  plot_df <- plot_df %>%
    dplyr::mutate(Genome_grp = factor(Genome_grp, levels = genome_levels))

  n_keep <- length(keep_genomes)

  if (is.null(genome_palette)) {
    pal_fun <- grDevices::colorRampPalette(c("steelblue4", "goldenrod2", "firebrick3"))
    keep_cols <- stats::setNames(pal_fun(n_keep), keep_genomes)
  } else {
    if (length(genome_palette) < n_keep) {
      stop("Not enough colours supplied in genome_palette for top_n genomes.")
    }
    keep_cols <- stats::setNames(genome_palette[seq_len(n_keep)], keep_genomes)
  }

  mag_cols <- c(keep_cols, Other = other_colour)

  legend_labels <- stats::setNames(genome_levels, genome_levels)

  if (!is.null(mags_tax)) {
    ab_map2 <- mags_tax %>%
      dplyr::select(
        Genome = dplyr::all_of(name_col),
        rel_abun = dplyr::all_of(rel_abun_col)
      ) %>%
      dplyr::filter(!is.na(Genome)) %>%
      dplyr::mutate(rel_abun = as.numeric(rel_abun)) %>%
      dplyr::distinct(Genome, .keep_all = TRUE)

    ab_vec <- ab_map2$rel_abun
    names(ab_vec) <- ab_map2$Genome

    keep_labels <- vapply(
      keep_genomes,
      function(g) {
        ra <- ab_vec[[g]]
        if (is.null(ra) || is.na(ra)) {
          sprintf("%s (NA%%)", g)
        } else {
          sprintf(paste0("%s (%.", rel_abun_digits, "f%%)"), g, ra)
        }
      },
      character(1)
    )
    names(keep_labels) <- keep_genomes

    legend_labels <- c(keep_labels, Other = "Other")
  }

  yvar <- if (mode == "proportion") "prop" else "count"
  ylab <- if (mode == "proportion") "Proportion of total gene count"
  else "Total gene count (summed across selected MAGs)"

  p <- ggplot2::ggplot(plot_df, ggplot2::aes(x = gene_id, y = .data[[yvar]], fill = Genome_grp)) +
    ggplot2::geom_col(width = 0.9) +
    ggplot2::scale_fill_manual(
      values = mag_cols,
      breaks = genome_levels,
      labels = legend_labels,
      drop = FALSE
    ) +
    ggplot2::labs(
      x = "CAZy family (gene_id)",
      y = ylab,
      title = title,
      fill = "MAG / species"
    ) +
    ggplot2::theme_bw() +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, vjust = 1, size = 7),
      legend.position = "right"
    )

  if (facet_by_class && !is.null(gene_classes)) {
    p <- p +
      ggplot2::facet_grid(. ~ class, scales = "free_x", space = "free_x", switch = "x") +
      ggplot2::theme(
        panel.spacing.x = grid::unit(0.8, "lines"),
        strip.placement = "outside",
        strip.background = ggplot2::element_rect(fill = "grey95", colour = NA)
      )
  }

  p <- p +
    ggplot2::scale_y_continuous(
      labels = if (mode == "proportion") scales::percent_format(accuracy = 1) else waiver(),
      expand = ggplot2::expansion(mult = c(0, 0.05))
    ) +
    ggplot2::coord_cartesian(ylim = c(0, 45))

  return(p)
}

p_meso_stack <- make_stacked_contrib_plot(
  df_counts = gene_count_meso_pectin,
  gene_classes = gene_classes,
  top_n = 10,
  mode = "count",
  mags_tax = meso_mags_tax,
  name_col = "new",
  rel_abun_col = "rel_abun",
  rel_abun_digits = 2,
  genome_palette = c(
    "orchid1", "purple", "slateblue", "slateblue4", "navy",
    "steelblue3", "mediumaquamarine",
    "seagreen1", "seagreen3", "olivedrab"
  ),
  other_colour = "grey80",
  title = "Mesophilic contribution to pectin CAZy gene counts"
)

p_thermo_stack <- make_stacked_contrib_plot(
  df_counts = gene_count_thermo_pectin,
  gene_classes = gene_classes,
  top_n = 10,
  mode = "count",
  mags_tax = thermo_mags_tax,
  name_col = "new",
  rel_abun_col = "rel_abun",
  rel_abun_digits = 2,
  genome_palette = c(
    "black", "firebrick4", "firebrick3",
    "firebrick1", "gold3", "gold1", "darkorange1",
    "chocolate", "darkorange4", "darkred"
  ),
  other_colour = "grey80",
  title = "Thermophilic contribution to pectin CAZy gene counts"
)

p <- p_meso_stack / p_thermo_stack

ggsave(filename = "pectin/only_abun_attribution_barplot.png", plot = p,
       width = 20, height = 12, dpi = 300)

```


Masked abundance stacked barplot
```{r}


make_abun_stack_plot <- function(df_counts,
                                 gene_classes,
                                 mags_tax,
                                 name_col = "new",
                                 rel_abun_col = "rel_abun",
                                 rel_abun_digits = 2,
                                 top_n = 10,
                                 title = "",
                                 y_lim = c(0, 45),
                                 other_colour = "grey80") {

  long <- df_counts %>%
    mutate(across(-gene_id, ~ as.numeric(.x))) %>%
    pivot_longer(-gene_id, names_to = "Genome", values_to = "count") %>%
    mutate(count = replace_na(count, 0))

  ab_map <- mags_tax %>%
    select(Genome = all_of(name_col),
           rel_abun = all_of(rel_abun_col)) %>%
    filter(!is.na(Genome)) %>%
    mutate(rel_abun = as.numeric(rel_abun)) %>%
    distinct(Genome, .keep_all = TRUE)

  long <- long %>% left_join(ab_map, by = "Genome")

  plot_df0 <- long %>%
    filter(count >= 1) %>%
    select(gene_id, Genome, rel_abun)

  genomes_with_any_gene <- plot_df0 %>%
    distinct(Genome) %>%
    pull(Genome)

  top_genomes <- ab_map %>%
    filter(Genome %in% genomes_with_any_gene) %>%
    arrange(desc(rel_abun)) %>%
    slice_head(n = top_n) %>%
    pull(Genome)

  plot_df <- plot_df0 %>%
    mutate(Genome_grp = if_else(Genome %in% top_genomes, Genome, "Other")) %>%
    group_by(gene_id, Genome_grp) %>%
    summarise(rel_abun = sum(rel_abun, na.rm = TRUE), .groups = "drop")

  gene_classes2 <- gene_classes %>%
    mutate(gene_id = as.character(gene_id),
           class = str_squish(class))

  gene_order <- gene_classes2 %>% pull(gene_id)
  class_levels <- gene_classes2 %>% distinct(class) %>% pull(class)

  plot_df <- plot_df %>%
    mutate(gene_id = as.character(gene_id)) %>%
    left_join(gene_classes2, by = "gene_id", relationship = "many-to-many") %>%
    mutate(
      class   = factor(class, levels = class_levels),
      gene_id = factor(gene_id, levels = unique(gene_order))
    )

  genome_levels <- c(top_genomes, "Other")
  plot_df <- plot_df %>%
    mutate(Genome_grp = factor(Genome_grp, levels = genome_levels))

  keep_cols <- setNames(scico::scico(length(top_genomes), palette = "managua"), top_genomes)
  mag_cols <- c(keep_cols, Other = other_colour)

  ab_vec <- ab_map$rel_abun
  names(ab_vec) <- ab_map$Genome

  keep_labels <- vapply(
    top_genomes,
    function(g) {
      ra <- ab_vec[[g]]
      if (is.null(ra) || is.na(ra)) sprintf("%s (NA%%)", g)
      else sprintf(paste0("%s (%.", rel_abun_digits, "f%%)"), g, ra)
    },
    character(1)
  )
  names(keep_labels) <- top_genomes
  legend_labels <- c(keep_labels, Other = "Other")

  p <- ggplot(plot_df, aes(x = gene_id, y = rel_abun, fill = Genome_grp)) +
    geom_col(width = 0.9) +
    scale_fill_manual(values = mag_cols, breaks = genome_levels, labels = legend_labels, drop = FALSE) +
    labs(x = "CAZy family (gene_id)", y = "Relative abundance (%)", title = title, fill = "MAG / species") +
    theme_bw() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 7),
      legend.position = "right"
    ) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
    coord_cartesian(ylim = y_lim) +
    facet_grid(. ~ class, scales = "free_x", space = "free_x", switch = "x") +
    theme(
      panel.spacing.x = unit(0.8, "lines"),
      strip.placement = "outside",
      strip.background = element_rect(fill = "grey95", colour = NA)
    )

  p
}

p_meso_stack <- make_abun_stack_plot(
  df_counts = gene_count_meso_pectin,
  gene_classes = gene_classes,
  mags_tax = meso_mags_tax,
  name_col = "new",
  rel_abun_col = "rel_abun",
  rel_abun_digits = 2,
  top_n = 10,
  y_lim = c(0, 14),
  title = "Mesophilic: relative abundance of genomes with CAZy count ≥ 1 (top 10 coloured)"
)

p_thermo_stack <- make_abun_stack_plot(
  df_counts = gene_count_thermo_pectin,
  gene_classes = gene_classes,
  mags_tax = thermo_mags_tax,
  name_col = "new",
  rel_abun_col = "rel_abun",
  rel_abun_digits = 2,
  top_n = 10,
  y_lim = c(0, 14),
  title = "Thermophilic: relative abundance of genomes with CAZy count ≥ 1 (top 10 coloured)"
)

p <- p_meso_stack / p_thermo_stack

ggsave("pectin/abun_masked_attribution_barplot.png",
       p, width = 20, height = 12, dpi = 300)

```


        --------------------------------------------------------------
    -----------------------------------------------------------------------
--------------------------------------------------------------------------------
                                     
                                    Cellulose

--------------------------------------------------------------------------------
    -----------------------------------------------------------------------
        --------------------------------------------------------------



--------------------------------------------------------------------------------
                                       Pre
--------------------------------------------------------------------------------

Important note regarding gene_id selection:
Only gene_ids for genes exclusively performing a specific function are selected. Others that perform multiple other functions, including the target function, were not selected due to uncertainty and a lack of interpretability. This means that we are potentially missing out on relevant gene_ids. 


Classifications (conservative)
```{r}

# Make classifications

gene_classes_cel_con <- tibble(
  gene_id = c(
    # Amorphous cellulose backbone cleavage
    "GH124", "GH6",

    # Amorphous cellulose oligo cleavage
    "GH94", "GH116",
    
    # Crystalline cellulose backbone cleavage
    "AA16", "AA11", "AA15", "AA10", "AA13"
  ),
  class = c(
    rep("Amorphous cellulose backbone cleavage", 2),
    rep("Amorphous cellulose oligo cleavage", 2),
    rep("Crystalline cellulose backbone cleavage", 5)
  )
)

```


Classifications (lenient)
```{r}

gene_classes_cel_len <- tibble(
  gene_id = c(
    # Amorphous cellulose backbone cleavage
    "GH124", "GH6", "GH9", "GH5", "GH8", "GH12", "GH44", "GH45", "GH10",

    # Amorphous cellulose oligo cleavage
    "GH94", "GH116", "GH1", "GH3", "GH16", "GH2",
    
    # Crystalline cellulose backbone cleavage
    "AA16","GH7", "AA11", "AA15", "AA10", "AA13", "AA9", "GH48"
  ),
  class = c(
    rep("Amorphous cellulose backbone cleavage", 9),
    rep("Amorphous cellulose oligo cleavage", 6),
    rep("Crystalline cellulose backbone cleavage", 8)
  )
)

```


Obtain target gene_ids for target MAGs
```{r}

# Filter for cellulose CAZYs (lenient)
gene_count_meso_cel_len <- gene_counts_meso %>%
  filter(gene_id %in% gene_classes_cel_len$gene_id)

gene_count_thermo_cel_len <- gene_counts_thermo %>%
  filter(gene_id %in% gene_classes_cel_len$gene_id)


# Filter for cellulose CAZYs (conservative)
gene_count_meso_cel_con <- gene_counts_meso %>%
  filter(gene_id %in% gene_classes_cel_con$gene_id)

gene_count_thermo_cel_con <- gene_counts_thermo %>%
  filter(gene_id %in% gene_classes_cel_con$gene_id)



```


--------------------------------------------------------------------------------
                                     Heatmap
--------------------------------------------------------------------------------

Note regarding the (lenient) heatmaps: The most prominent CAZy genes, GH2 and GH3 are multi-functional, encoding among others pectin deg and arabinan deg, making their interpretation for the relevance in cellulose deg impossible. They are perhaps not really involved in cellulose deg.


conservative
```{r}



meso_long <- gene_count_meso_cel_con %>%
  pivot_longer(
    cols = -gene_id,
    names_to = "genome",
    values_to = "count"
  )


thermo_long <- gene_count_thermo_cel_con %>%
  pivot_longer(
    cols = -gene_id,
    names_to = "genome",
    values_to = "count"
  )


p_meso <- ggplot(meso_long, aes(x = genome, y = gene_id, fill = count)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Count", limits = c(0, 15)) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10),
    axis.title = element_blank()
  ) +
  ggtitle("Cellulose CAZy repertoire – Mesophilic MAGs")


p_thermo <- ggplot(thermo_long, aes(x = genome, y = gene_id, fill = count)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Count", limits = c(0, 8)) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10),
    axis.title = element_blank()
  ) +
  ggtitle("Cellulose CAZy repertoire – Thermophilic MAGs")

p <- p_meso / p_thermo

ggsave(filename = "cellulose/gene_count_heatmap_conservative.png", plot = p,
       width = 20, height = 12, dpi = 300)

```


Lenient
```{r}



meso_long <- gene_count_meso_cel_len %>%
  pivot_longer(
    cols = -gene_id,
    names_to = "genome",
    values_to = "count"
  )


thermo_long <- gene_count_thermo_cel_len %>%
  pivot_longer(
    cols = -gene_id,
    names_to = "genome",
    values_to = "count"
  )


p_meso <- ggplot(meso_long, aes(x = genome, y = gene_id, fill = count)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Count", limits = c(0, 15)) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10),
    axis.title = element_blank()
  ) +
  ggtitle("Pectin CAZy repertoire – Mesophilic MAGs")


p_thermo <- ggplot(thermo_long, aes(x = genome, y = gene_id, fill = count)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Count", limits = c(0, 8)) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10),
    axis.title = element_blank()
  ) +
  ggtitle("Pectin CAZy repertoire – Thermophilic MAGs")

p <- p_meso / p_thermo

ggsave(filename = "cellulose/gene_count_heatmap_lenient.png", plot = p,
       width = 20, height = 12, dpi = 300)

```


--------------------------------------------------------------------------------
                                     Barplots
--------------------------------------------------------------------------------


Class-separated
```{r}
### Lenient

library(tidyverse)
library(scales)
library(grid)  # unit()

# ------------------------------------------------------------
# Helper: count MAGs with count > 0 per gene_id
# ------------------------------------------------------------
count_mags_per_gene <- function(df_counts) {
  df_counts %>%
    mutate(across(-gene_id, ~ as.numeric(.x))) %>%
    pivot_longer(-gene_id, names_to = "Genome", values_to = "count") %>%
    mutate(present = !is.na(count) & count > 0) %>%
    group_by(gene_id) %>%
    summarise(n_MAGs = sum(present), .groups = "drop")
}

# helper: make labels fit in skinny facets
stack_gene_label <- function(x) {
  x <- as.character(x)
  gsub("^([A-Z]+)([0-9]+)$", "\\1\n\\2", x)
}

# ------------------------------------------------------------
# Plot function: facet by class + class colours from a gradient
# ------------------------------------------------------------
plot_cazy_by_class <- function(df_counts, gene_classes_cel_len, title, y_max = NULL,
                               gradient_anchors = c("grey80", "steelblue4"),
                               label_stack = TRUE) {

  hits <- count_mags_per_gene(df_counts)

  gene_order <- gene_classes_cel_len %>%
    mutate(gene_id = as.character(gene_id)) %>%
    pull(gene_id)

  plot_df <- gene_classes_cel_len %>%
    mutate(
      gene_id = as.character(gene_id),
      class   = str_squish(class)
    ) %>%
    left_join(hits, by = "gene_id") %>%
    mutate(
      n_MAGs  = replace_na(n_MAGs, 0L),
      class   = fct_inorder(class),
      gene_id = factor(gene_id, levels = unique(gene_order))
    )

  class_levels <- gene_classes_cel_len %>%
    mutate(class = str_squish(class)) %>%
    distinct(class) %>%
    pull(class)

  plot_df <- plot_df %>%
    mutate(class = factor(as.character(class), levels = class_levels))

  pal_fun <- colorRampPalette(gradient_anchors)
  class_cols <- setNames(pal_fun(length(class_levels)), class_levels)

  p <- ggplot(plot_df, aes(x = gene_id, y = n_MAGs, fill = class)) +
    geom_col(width = 0.9) +
    facet_grid(. ~ class, scales = "free_x", space = "free_x", switch = "x") +
    scale_fill_manual(values = class_cols, drop = FALSE) +
    geom_text(aes(label = n_MAGs), vjust = -0.3, size = 2.5) +
    labs(
      x = NULL,
      y = "Number of MAGs (gene count > 0)",
      title = title,
      fill = "Class"
    ) +
    theme_bw() +
    theme(
      axis.text.x = element_text(
        angle = 0,
        hjust = 0.5, vjust = 1,
        size = 7,
        lineheight = 0.9
      ),
      axis.ticks.x = element_line(linewidth = 0.25),
      panel.spacing.x = unit(0.9, "lines"),
      legend.position = "none",
      strip.placement = "outside",
      strip.background = element_rect(fill = "grey95", colour = NA),
      plot.margin = margin(t = 5.5, r = 5.5, b = 18, l = 5.5)
    ) +
    scale_y_continuous(
      limit = c(0, 50),
      expand = expansion(mult = c(0, 0.10))
    ) +
    coord_cartesian(clip = "off")

  if (isTRUE(label_stack)) {
    p <- p + scale_x_discrete(labels = stack_gene_label)
  } else {
    p <- p + scale_x_discrete(drop = FALSE)
  }

  p
}

# ------------------------------------------------------------
# Run for meso (blue gradient) and thermo (red gradient)
# using gene_classes_cel_len
# ------------------------------------------------------------
p_meso <- plot_cazy_by_class(
  df_counts = gene_count_meso_cel_len,        # <-- your cellulose (len) count table
  gene_classes_cel_len = gene_classes_cel_len,
  title = "Number of mesophilic MAGs encoding each cellulose CAZy (count > 0)",
  y_max = 20,
  gradient_anchors = c("cornflowerblue", "royalblue2", "royalblue3", "royalblue4", "midnightblue"),
  label_stack = TRUE
)

p_thermo <- plot_cazy_by_class(
  df_counts = gene_count_thermo_cel_len,      # <-- your cellulose (len) count table
  gene_classes_cel_len = gene_classes_cel_len,
  title = "Number of thermophilic MAGs encoding each cellulose CAZy (count > 0)",
  y_max = 20,
  gradient_anchors = c("sandybrown", "sienna1", "sienna2", "sienna3", "sienna", "sienna4"),
  label_stack = TRUE
)

p <- p_meso / p_thermo
p



ggsave(filename = "cellulose/mag_gene_counts_barplot_lenient.png", plot = p,
       width = 20, height = 12, dpi = 300)





### Conservative

# ------------------------------------------------------------
# Helper: count MAGs with count > 0 per gene_id
# ------------------------------------------------------------
count_mags_per_gene <- function(df_counts) {
  df_counts %>%
    mutate(across(-gene_id, ~ as.numeric(.x))) %>%
    pivot_longer(-gene_id, names_to = "Genome", values_to = "count") %>%
    mutate(present = !is.na(count) & count > 0) %>%
    group_by(gene_id) %>%
    summarise(n_MAGs = sum(present), .groups = "drop")
}

# helper: make labels fit in skinny facets
stack_gene_label <- function(x) {
  x <- as.character(x)
  gsub("^([A-Z]+)([0-9]+)$", "\\1\n\\2", x)
}

# ------------------------------------------------------------
# Plot function: facet by class + class colours from a gradient
# ------------------------------------------------------------
plot_cazy_by_class <- function(df_counts, gene_classes_cel_con, title, y_max = NULL,
                               gradient_anchors = c("grey80", "steelblue4"),
                               label_stack = TRUE) {

  hits <- count_mags_per_gene(df_counts)

  gene_order <- gene_classes_cel_con %>%
    mutate(gene_id = as.character(gene_id)) %>%
    pull(gene_id)

  plot_df <- gene_classes_cel_con %>%
    mutate(
      gene_id = as.character(gene_id),
      class   = str_squish(class)
    ) %>%
    left_join(hits, by = "gene_id") %>%
    mutate(
      n_MAGs  = replace_na(n_MAGs, 0L),
      class   = fct_inorder(class),
      gene_id = factor(gene_id, levels = unique(gene_order))
    )

  class_levels <- gene_classes_cel_con %>%
    mutate(class = str_squish(class)) %>%
    distinct(class) %>%
    pull(class)

  plot_df <- plot_df %>%
    mutate(class = factor(as.character(class), levels = class_levels))

  pal_fun <- colorRampPalette(gradient_anchors)
  class_cols <- setNames(pal_fun(length(class_levels)), class_levels)

  p <- ggplot(plot_df, aes(x = gene_id, y = n_MAGs, fill = class)) +
    geom_col(width = 0.9) +
    facet_grid(. ~ class, scales = "free_x", space = "free_x", switch = "x") +
    scale_fill_manual(values = class_cols, drop = FALSE) +
    geom_text(aes(label = n_MAGs), vjust = -0.3, size = 2.5) +
    labs(
      x = NULL,
      y = "Number of MAGs (gene count > 0)",
      title = title,
      fill = "Class"
    ) +
    theme_bw() +
    theme(
      axis.text.x = element_text(
        angle = 0,
        hjust = 0.5, vjust = 1,
        size = 7,
        lineheight = 0.9
      ),
      axis.ticks.x = element_line(linewidth = 0.25),
      panel.spacing.x = unit(0.9, "lines"),
      legend.position = "none",
      strip.placement = "outside",
      strip.background = element_rect(fill = "grey95", colour = NA),
      plot.margin = margin(t = 5.5, r = 5.5, b = 18, l = 5.5)
    ) +
    scale_y_continuous(
      limit = c(0, 50),
      expand = expansion(mult = c(0, 0.10))
    ) +
    coord_cartesian(clip = "off")

  if (isTRUE(label_stack)) {
    p <- p + scale_x_discrete(labels = stack_gene_label)
  } else {
    p <- p + scale_x_discrete(drop = FALSE)
  }

  p
}

# ------------------------------------------------------------
# Run for meso (blue gradient) and thermo (red gradient)
# using gene_classes_cel_len
# ------------------------------------------------------------
p_meso <- plot_cazy_by_class(
  df_counts = gene_count_meso_cel_con,
  gene_classes_cel_con = gene_classes_cel_con,
  title = "Number of mesophilic MAGs encoding each cellulose CAZy (count > 0) (conservative)",
  y_max = 20,
  gradient_anchors = c("cornflowerblue", "royalblue2", "royalblue3", "royalblue4", "midnightblue"),
  label_stack = TRUE
)

p_thermo <- plot_cazy_by_class(
  df_counts = gene_count_thermo_cel_con,      # <-- your cellulose (len) count table
  gene_classes_cel_con = gene_classes_cel_con,
  title = "Number of thermophilic MAGs encoding each cellulose CAZy (count > 0) (conservative)",
  y_max = 20,
  gradient_anchors = c("sandybrown", "sienna1", "sienna2", "sienna3", "sienna", "sienna4"),
  label_stack = TRUE
)

p <- p_meso / p_thermo
p



ggsave(filename = "cellulose/mag_gene_counts_barplot_conservative.png", plot = p,
       width = 20, height = 12, dpi = 300)



```


Attribution plot
```{r}


make_stacked_contrib_plot <- function(df_counts,
                                      gene_classes,
                                      top_n = 10,
                                      title = "",
                                      mode = c("proportion", "count"),
                                      facet_by_class = TRUE,
                                      genome_palette = NULL,
                                      other_colour = "grey70",
                                      y_lim = NULL,
                                      mags_tax = NULL,
                                      name_col = "new",
                                      rel_abun_col = "rel_abun",
                                      rel_abun_digits = 2,
                                      select_top_by = c("gene_count", "abundance")) {

  mode <- match.arg(mode)
  select_top_by <- match.arg(select_top_by)

  long <- df_counts %>%
    mutate(across(-gene_id, ~ as.numeric(.x))) %>%
    pivot_longer(-gene_id, names_to = "Genome", values_to = "count") %>%
    mutate(count = replace_na(count, 0))

  present_genomes <- unique(long$Genome)

  if (select_top_by == "abundance") {
    if (is.null(mags_tax)) stop("select_top_by='abundance' requires mags_tax.")

    ab_map <- mags_tax %>%
      select(Genome = all_of(name_col), rel_abun = all_of(rel_abun_col)) %>%
      filter(!is.na(Genome)) %>%
      mutate(rel_abun = as.numeric(rel_abun)) %>%
      distinct(Genome, .keep_all = TRUE)

    keep_genomes <- ab_map %>%
      filter(Genome %in% present_genomes) %>%
      arrange(desc(rel_abun)) %>%
      slice_head(n = top_n) %>%
      pull(Genome)

  } else {
    keep_genomes <- long %>%
      group_by(Genome) %>%
      summarise(total = sum(count), .groups = "drop") %>%
      arrange(desc(total)) %>%
      slice_head(n = top_n) %>%
      pull(Genome)
  }

  long2 <- long %>%
    mutate(Genome_grp = if_else(Genome %in% keep_genomes, Genome, "Other"))

  plot_df <- long2 %>%
    group_by(gene_id, Genome_grp) %>%
    summarise(count = sum(count), .groups = "drop")

  if (mode == "proportion") {
    plot_df <- plot_df %>%
      group_by(gene_id) %>%
      mutate(denom = sum(count), prop = if_else(denom > 0, count / denom, 0)) %>%
      ungroup() %>%
      select(-denom)
  }

  gene_order <- gene_classes %>% mutate(gene_id = as.character(gene_id)) %>% pull(gene_id)
  class_levels <- gene_classes %>% mutate(class = str_squish(class)) %>% distinct(class) %>% pull(class)

  plot_df <- plot_df %>%
    mutate(gene_id = as.character(gene_id)) %>%
    left_join(gene_classes %>% mutate(gene_id = as.character(gene_id), class = str_squish(class)),
              by = "gene_id") %>%
    mutate(
      class   = factor(class, levels = class_levels),
      gene_id = factor(gene_id, levels = unique(gene_order))
    )

  genome_levels <- c(keep_genomes, "Other")
  plot_df <- plot_df %>% mutate(Genome_grp = factor(Genome_grp, levels = genome_levels))

  n_keep <- length(keep_genomes)

  if (is.null(genome_palette)) {
    pal_fun <- grDevices::colorRampPalette(c("steelblue4", "goldenrod2", "firebrick3"))
    keep_cols <- stats::setNames(pal_fun(n_keep), keep_genomes)
  } else {
    if (length(genome_palette) < n_keep) stop("Not enough colours supplied in genome_palette.")
    keep_cols <- stats::setNames(genome_palette[seq_len(n_keep)], keep_genomes)
  }

  mag_cols <- c(keep_cols, Other = other_colour)

  legend_labels <- stats::setNames(genome_levels, genome_levels)

  if (!is.null(mags_tax)) {
    ab_map2 <- mags_tax %>%
      select(Genome = all_of(name_col), rel_abun = all_of(rel_abun_col)) %>%
      filter(!is.na(Genome)) %>%
      mutate(rel_abun = as.numeric(rel_abun)) %>%
      distinct(Genome, .keep_all = TRUE)

    ab_vec <- ab_map2$rel_abun
    names(ab_vec) <- ab_map2$Genome

    keep_labels <- vapply(
      keep_genomes,
      function(g) {
        ra <- ab_vec[[g]]
        if (is.null(ra) || is.na(ra)) sprintf("%s (NA%%)", g)
        else sprintf(paste0("%s (%.", rel_abun_digits, "f%%)"), g, ra)
      },
      character(1)
    )
    names(keep_labels) <- keep_genomes
    legend_labels <- c(keep_labels, Other = "Other")
  }

  yvar <- if (mode == "proportion") "prop" else "count"
  ylab <- if (mode == "proportion") "Proportion of total gene count"
  else "Total gene count (summed across selected MAGs)"

  p <- ggplot(plot_df, aes(x = gene_id, y = .data[[yvar]], fill = Genome_grp)) +
    geom_col(width = 0.9) +
    scale_fill_manual(values = mag_cols, breaks = genome_levels, labels = legend_labels, drop = FALSE) +
    labs(x = "CAZy family (gene_id)", y = ylab, title = title, fill = "MAG / species") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 7),
          legend.position = "right")

  if (facet_by_class) {
    p <- p +
      facet_grid(. ~ class, scales = "free_x", space = "free_x", switch = "x") +
      theme(panel.spacing.x = unit(0.8, "lines"),
            strip.placement = "outside",
            strip.background = element_rect(fill = "grey95", colour = NA))
  }

  p <- p +
    scale_y_continuous(
      labels = if (mode == "proportion") percent_format(accuracy = 1) else waiver(),
      expand = expansion(mult = c(0, 0.05))
    )

  if (!is.null(y_lim)) p <- p + coord_cartesian(ylim = y_lim)
  p
}

GENE_CLASSES <- gene_classes_cel_len

# Use the exact sample-specific palettes you like (same approach as your working pectin code)
PAL_MESO_10 <- c(
  "orchid1", "purple", "slateblue", "slateblue4", "navy",
  "steelblue3", "mediumaquamarine", "seagreen1", "seagreen3", "olivedrab"
)

PAL_THERMO_10 <- c(
  "black", "firebrick4", "firebrick3",
  "firebrick1", "gold3", "gold1", "darkorange1",
  "chocolate", "darkorange4", "darkred"
)

# Top 10 by abundance (lenient)
p_meso_abun_len <- make_stacked_contrib_plot(
  df_counts = gene_count_meso_cel_len,
  gene_classes = GENE_CLASSES,
  top_n = 10,
  select_top_by = "abundance",
  mode = "count",
  mags_tax = meso_mags_tax,
  name_col = "new",
  rel_abun_col = "rel_abun",
  rel_abun_digits = 2,
  genome_palette = PAL_MESO_10,
  other_colour = "grey80",
  y_lim = c(0, 130),
  title = "Mesophilic contribution (cellulose CAZy gene counts) (lenient) — top 10 by abundance"
)

p_thermo_abun_len <- make_stacked_contrib_plot(
  df_counts = gene_count_thermo_cel_len,
  gene_classes = GENE_CLASSES,
  top_n = 10,
  select_top_by = "abundance",
  mode = "count",
  mags_tax = thermo_mags_tax,
  name_col = "new",
  rel_abun_col = "rel_abun",
  rel_abun_digits = 2,
  genome_palette = PAL_THERMO_10,
  other_colour = "grey80",
  y_lim = c(0, 130),
  title = "Thermophilic contribution (cellulose CAZy gene counts) (lenient) — top 10 by abundance"
)

p_abun_len <- p_meso_abun_len / p_thermo_abun_len

ggsave("cellulose/abun_attribution_barplot_lenient_top10.png", p_abun_len,
       width = 20, height = 12, dpi = 300)

# Top 10 by abundance (conservative)
p_meso_abun_con <- make_stacked_contrib_plot(
  df_counts = gene_count_meso_cel_con,
  gene_classes = GENE_CLASSES,
  top_n = 10,
  select_top_by = "abundance",
  mode = "count",
  mags_tax = meso_mags_tax,
  name_col = "new",
  rel_abun_col = "rel_abun",
  rel_abun_digits = 2,
  genome_palette = PAL_MESO_10,
  other_colour = "grey80",
  y_lim = c(0, 25),
  title = "Mesophilic contribution (cellulose CAZy gene counts) (conservative) — top 10 by abundance"
)

p_thermo_abun_con <- make_stacked_contrib_plot(
  df_counts = gene_count_thermo_cel_con,
  gene_classes = GENE_CLASSES,
  top_n = 10,
  select_top_by = "abundance",
  mode = "count",
  mags_tax = thermo_mags_tax,
  name_col = "new",
  rel_abun_col = "rel_abun",
  rel_abun_digits = 2,
  genome_palette = PAL_THERMO_10,
  other_colour = "grey80",
  y_lim = c(0, 25),
  title = "Thermophilic contribution (cellulose CAZy gene counts) (conservative) — top 10 by abundance"
)

p_abun_con <- p_meso_abun_con / p_thermo_abun_con

ggsave("cellulose/abun_attribution_barplot_conservative_top10.png", p_abun_con,
       width = 20, height = 12, dpi = 300)

ggsave(
  filename = "cellulose/abun_attribution_barplot_conservative_top10.png",
  plot = p_abun_con,
  width = 20, height = 12, dpi = 300
)

```


Abundance only attribute plot
```{r}

make_stacked_contrib_plot <- function(df_counts,
                                      gene_classes,
                                      top_n = 10,
                                      title = "",
                                      mode = c("proportion", "count"),
                                      facet_by_class = TRUE,
                                      genome_palette = NULL,
                                      other_colour = "grey70",
                                      y_lim = NULL,
                                      mags_tax = NULL,
                                      name_col = "new",
                                      rel_abun_col = "rel_abun",
                                      rel_abun_digits = 2,
                                      select_top_by = c("gene_count", "abundance")) {

  mode <- match.arg(mode)
  select_top_by <- match.arg(select_top_by)

  long <- df_counts %>%
    mutate(across(-gene_id, ~ as.numeric(.x))) %>%
    pivot_longer(-gene_id, names_to = "Genome", values_to = "count") %>%
    mutate(count = replace_na(count, 0))

  present_genomes <- unique(long$Genome)

  if (select_top_by == "abundance") {
    if (is.null(mags_tax)) stop("select_top_by='abundance' requires mags_tax.")

    ab_map <- mags_tax %>%
      select(Genome = all_of(name_col), rel_abun = all_of(rel_abun_col)) %>%
      filter(!is.na(Genome)) %>%
      mutate(rel_abun = as.numeric(rel_abun)) %>%
      distinct(Genome, .keep_all = TRUE)

    genomes_with_any_gene <- long %>%
      group_by(Genome) %>%
      summarise(has_gene = any(count >= 1), .groups = "drop") %>%
      filter(has_gene) %>%
      pull(Genome)

    keep_genomes <- ab_map %>%
      filter(Genome %in% present_genomes, Genome %in% genomes_with_any_gene) %>%
      arrange(desc(rel_abun)) %>%
      slice_head(n = top_n) %>%
      pull(Genome)

  } else {
    keep_genomes <- long %>%
      group_by(Genome) %>%
      summarise(total = sum(count), .groups = "drop") %>%
      arrange(desc(total)) %>%
      slice_head(n = top_n) %>%
      pull(Genome)
  }

  long2 <- long %>%
    mutate(Genome_grp = if_else(Genome %in% keep_genomes, Genome, "Other"))

  plot_df <- long2 %>%
    group_by(gene_id, Genome_grp) %>%
    summarise(count = sum(count), .groups = "drop")

  if (mode == "proportion") {
    plot_df <- plot_df %>%
      group_by(gene_id) %>%
      mutate(denom = sum(count), prop = if_else(denom > 0, count / denom, 0)) %>%
      ungroup() %>%
      select(-denom)
  }

  gene_order <- gene_classes %>% mutate(gene_id = as.character(gene_id)) %>% pull(gene_id)
  class_levels <- gene_classes %>% mutate(class = str_squish(class)) %>% distinct(class) %>% pull(class)

  plot_df <- plot_df %>%
    mutate(gene_id = as.character(gene_id)) %>%
    left_join(gene_classes %>% mutate(gene_id = as.character(gene_id), class = str_squish(class)),
              by = "gene_id") %>%
    mutate(
      class   = factor(class, levels = class_levels),
      gene_id = factor(gene_id, levels = unique(gene_order))
    )

  genome_levels <- c(keep_genomes, "Other")
  plot_df <- plot_df %>% mutate(Genome_grp = factor(Genome_grp, levels = genome_levels))

  n_keep <- length(keep_genomes)

  if (is.null(genome_palette)) {
    pal_fun <- grDevices::colorRampPalette(c("steelblue4", "goldenrod2", "firebrick3"))
    keep_cols <- stats::setNames(pal_fun(n_keep), keep_genomes)
  } else {
    if (length(genome_palette) < n_keep) stop("Not enough colours supplied in genome_palette.")
    keep_cols <- stats::setNames(genome_palette[seq_len(n_keep)], keep_genomes)
  }

  mag_cols <- c(keep_cols, Other = other_colour)

  legend_labels <- stats::setNames(genome_levels, genome_levels)

  if (!is.null(mags_tax)) {
    ab_map2 <- mags_tax %>%
      select(Genome = all_of(name_col), rel_abun = all_of(rel_abun_col)) %>%
      filter(!is.na(Genome)) %>%
      mutate(rel_abun = as.numeric(rel_abun)) %>%
      distinct(Genome, .keep_all = TRUE)

    ab_vec <- ab_map2$rel_abun
    names(ab_vec) <- ab_map2$Genome

    keep_labels <- vapply(
      keep_genomes,
      function(g) {
        ra <- ab_vec[[g]]
        if (is.null(ra) || is.na(ra)) sprintf("%s (NA%%)", g)
        else sprintf(paste0("%s (%.", rel_abun_digits, "f%%)"), g, ra)
      },
      character(1)
    )
    names(keep_labels) <- keep_genomes
    legend_labels <- c(keep_labels, Other = "Other")
  }

  yvar <- if (mode == "proportion") "prop" else "count"
  ylab <- if (mode == "proportion") "Proportion of total gene count"
  else "Total gene count (summed across selected MAGs)"

  p <- ggplot(plot_df, aes(x = gene_id, y = .data[[yvar]], fill = Genome_grp)) +
    geom_col(width = 0.9) +
    scale_fill_manual(values = mag_cols, breaks = genome_levels, labels = legend_labels, drop = FALSE) +
    labs(x = "CAZy family (gene_id)", y = ylab, title = title, fill = "MAG / species") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 7),
          legend.position = "right")

  if (facet_by_class) {
    p <- p +
      facet_grid(. ~ class, scales = "free_x", space = "free_x", switch = "x") +
      theme(panel.spacing.x = unit(0.8, "lines"),
            strip.placement = "outside",
            strip.background = element_rect(fill = "grey95", colour = NA))
  }

  p <- p +
    scale_y_continuous(
      labels = if (mode == "proportion") percent_format(accuracy = 1) else waiver(),
      expand = expansion(mult = c(0, 0.05))
    )

  if (!is.null(y_lim)) p <- p + coord_cartesian(ylim = y_lim)
  p
}

GENE_CLASSES <- gene_classes_cel_len

PAL_MESO_10 <- c(
  "orchid1", "purple", "slateblue", "slateblue4", "navy",
  "steelblue3", "mediumaquamarine", "seagreen1", "seagreen3", "olivedrab"
)

PAL_THERMO_10 <- c(
  "black", "firebrick4", "firebrick3",
  "firebrick1", "gold3", "gold1", "darkorange1",
  "chocolate", "darkorange4", "darkred"
)

p_meso_abun_len <- make_stacked_contrib_plot(
  df_counts = gene_count_meso_cel_len,
  gene_classes = GENE_CLASSES,
  top_n = 10,
  select_top_by = "abundance",
  mode = "count",
  mags_tax = meso_mags_tax,
  name_col = "new",
  rel_abun_col = "rel_abun",
  rel_abun_digits = 2,
  genome_palette = PAL_MESO_10,
  other_colour = "grey80",
  y_lim = c(0, 130),
  title = "Mesophilic contribution (cellulose CAZy gene counts) (lenient) — top 10 by abundance (>=1 gene)"
)

p_thermo_abun_len <- make_stacked_contrib_plot(
  df_counts = gene_count_thermo_cel_len,
  gene_classes = GENE_CLASSES,
  top_n = 10,
  select_top_by = "abundance",
  mode = "count",
  mags_tax = thermo_mags_tax,
  name_col = "new",
  rel_abun_col = "rel_abun",
  rel_abun_digits = 2,
  genome_palette = PAL_THERMO_10,
  other_colour = "grey80",
  y_lim = c(0, 130),
  title = "Thermophilic contribution (cellulose CAZy gene counts) (lenient) — top 10 by abundance (>=1 gene)"
)

p_abun_len <- p_meso_abun_len / p_thermo_abun_len

ggsave("cellulose/only_abun_attribution_barplot_lenient_top10.png", p_abun_len,
       width = 20, height = 12, dpi = 300)

p_meso_abun_con <- make_stacked_contrib_plot(
  df_counts = gene_count_meso_cel_con,
  gene_classes = GENE_CLASSES,
  top_n = 10,
  select_top_by = "abundance",
  mode = "count",
  mags_tax = meso_mags_tax,
  name_col = "new",
  rel_abun_col = "rel_abun",
  rel_abun_digits = 2,
  genome_palette = PAL_MESO_10,
  other_colour = "grey80",
  y_lim = c(0, 25),
  title = "Mesophilic contribution (cellulose CAZy gene counts) (conservative) — top 10 by abundance (>=1 gene)"
)

p_thermo_abun_con <- make_stacked_contrib_plot(
  df_counts = gene_count_thermo_cel_con,
  gene_classes = GENE_CLASSES,
  top_n = 10,
  select_top_by = "abundance",
  mode = "count",
  mags_tax = thermo_mags_tax,
  name_col = "new",
  rel_abun_col = "rel_abun",
  rel_abun_digits = 2,
  genome_palette = PAL_THERMO_10,
  other_colour = "grey80",
  y_lim = c(0, 25),
  title = "Thermophilic contribution (cellulose CAZy gene counts) (conservative) — top 10 by abundance (>=1 gene)"
)

p_abun_con <- p_meso_abun_con / p_thermo_abun_con

ggsave("cellulose/only_abun_attribution_barplot_conservative_top10.png", p_abun_con,
       width = 20, height = 12, dpi = 300)

```


Masked abundance stacked barplot
```{r}

make_abun_stack_plot <- function(df_counts,
                                 gene_classes,
                                 title = "",
                                 facet_by_class = TRUE,
                                 mags_tax,
                                 name_col = "new",
                                 rel_abun_col = "rel_abun",
                                 top_n_abun = 20,
                                 other_colour = "grey85",
                                 y_lim = NULL) {

  long <- df_counts %>%
    mutate(across(-gene_id, ~ as.numeric(.x))) %>%
    pivot_longer(-gene_id, names_to = "Genome", values_to = "count") %>%
    mutate(count = replace_na(count, 0))

  ab_map <- mags_tax %>%
    select(Genome = all_of(name_col), rel_abun = all_of(rel_abun_col)) %>%
    filter(!is.na(Genome)) %>%
    mutate(rel_abun = as.numeric(rel_abun)) %>%
    distinct(Genome, .keep_all = TRUE)

  long2 <- long %>%
    filter(count >= 1) %>%
    left_join(ab_map, by = "Genome")

  # keep top N by abundance among genomes that appear (count>=1 for >=1 gene)
  top_genomes <- long2 %>%
    distinct(Genome, rel_abun) %>%
    arrange(desc(rel_abun)) %>%
    slice_head(n = top_n_abun) %>%
    pull(Genome)

  long3 <- long2 %>%
    mutate(Genome_grp = if_else(Genome %in% top_genomes, Genome, "Other")) %>%
    group_by(gene_id, Genome_grp) %>%
    summarise(rel_abun = sum(rel_abun, na.rm = TRUE), .groups = "drop")

  gene_classes2 <- gene_classes %>%
    mutate(gene_id = as.character(gene_id), class = str_squish(class))

  gene_order <- gene_classes2 %>% pull(gene_id)
  class_levels <- gene_classes2 %>% distinct(class) %>% pull(class)

  plot_df <- long3 %>%
    mutate(gene_id = as.character(gene_id)) %>%
    left_join(gene_classes2, by = "gene_id", relationship = "many-to-many") %>%
    mutate(
      class   = factor(class, levels = class_levels),
      gene_id = factor(gene_id, levels = unique(gene_order))
    )

  genome_levels <- c(top_genomes, "Other")
  plot_df <- plot_df %>% mutate(Genome_grp = factor(Genome_grp, levels = genome_levels))

  # colours: viridis for top genomes, grey for Other
  keep_cols <- setNames(
  scico(length(top_genomes), palette = "managua"),
  top_genomes
)
  mag_cols <- c(keep_cols, Other = other_colour)

  # legend labels include abundance
  ab_vec <- ab_map$rel_abun
  names(ab_vec) <- ab_map$Genome

  legend_labels <- c(
    setNames(
      vapply(
        top_genomes,
        function(g) {
          ra <- ab_vec[[g]]
          if (is.null(ra) || is.na(ra)) sprintf("%s (NA%%)", g) else sprintf("%s (%.2f%%)", g, ra)
        },
        character(1)
      ),
      top_genomes
    ),
    Other = "Other"
  )

  p <- ggplot(plot_df, aes(x = gene_id, y = rel_abun, fill = Genome_grp)) +
    geom_col(width = 0.9) +
    scale_fill_manual(values = mag_cols, breaks = genome_levels, labels = legend_labels, drop = FALSE) +
    labs(
      x = "CAZy family (gene_id)",
      y = "Relative abundance (%) of genomes with count ≥ 1",
      title = title,
      fill = "MAG / species"
    ) +
    theme_bw() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 7),
      legend.position = "right"
    ) +
    scale_y_continuous(labels = function(x) paste0(x, "%"), expand = expansion(mult = c(0, 0.05)))

  if (facet_by_class) {
    p <- p +
      facet_grid(. ~ class, scales = "free_x", space = "free_x", switch = "x") +
      theme(
        panel.spacing.x = unit(0.8, "lines"),
        strip.placement = "outside",
        strip.background = element_rect(fill = "grey95", colour = NA)
      )
  }

  if (!is.null(y_lim)) p <- p + coord_cartesian(ylim = y_lim)
  p
}

GENE_CLASSES <- gene_classes_cel_len

# lenient
p_meso_abunmask_len <- make_abun_stack_plot(
  df_counts = gene_count_meso_cel_len,
  gene_classes = GENE_CLASSES,
  mags_tax = meso_mags_tax,
  name_col = "new",
  rel_abun_col = "rel_abun",
  top_n_abun = 20,
  title = "Mesophilic (lenient): abundance of genomes with CAZy count ≥ 1 (top 20 coloured)"
)

p_thermo_abunmask_len <- make_abun_stack_plot(
  df_counts = gene_count_thermo_cel_len,
  gene_classes = GENE_CLASSES,
  mags_tax = thermo_mags_tax,
  name_col = "new",
  rel_abun_col = "rel_abun",
  top_n_abun = 20,
  title = "Thermophilic (lenient): abundance of genomes with CAZy count ≥ 1 (top 20 coloured)"
)

p_len <- p_meso_abunmask_len / p_thermo_abunmask_len

ggsave("cellulose/abun_masked_by_gene_lenient_top20.png", p_len,
       width = 20, height = 12, dpi = 300)

# conservative
p_meso_abunmask_con <- make_abun_stack_plot(
  df_counts = gene_count_meso_cel_con,
  gene_classes = GENE_CLASSES,
  mags_tax = meso_mags_tax,
  name_col = "new",
  rel_abun_col = "rel_abun",
  top_n_abun = 20,
  title = "Mesophilic (conservative): abundance of genomes with CAZy count ≥ 1 (top 20 coloured)"
)

p_thermo_abunmask_con <- make_abun_stack_plot(
  df_counts = gene_count_thermo_cel_con,
  gene_classes = GENE_CLASSES,
  mags_tax = thermo_mags_tax,
  name_col = "new",
  rel_abun_col = "rel_abun",
  top_n_abun = 20,
  title = "Thermophilic (conservative): abundance of genomes with CAZy count ≥ 1 (top 20 coloured)"
)

p_con <- p_meso_abunmask_con | p_thermo_abunmask_con

ggsave("cellulose/abun_masked_by_gene_conservative_top20.png", p_con,
       width = 15, height = 7, dpi = 300)


```


        --------------------------------------------------------------
    -----------------------------------------------------------------------
--------------------------------------------------------------------------------
                                     
                                  Hemicellulose

--------------------------------------------------------------------------------
    -----------------------------------------------------------------------
        --------------------------------------------------------------



--------------------------------------------------------------------------------
                                       Pre
--------------------------------------------------------------------------------

Important note regarding gene_id selection:
Only gene_ids for genes exclusively performing a specific function are selected. Others that perform multiple other functions, including the target function, were not selected due to uncertainty and a lack of interpretability. This means that we are potentially missing out on relevant gene_ids. 


Classifications (conservative)
```{r}



gene_classes_hem_con <- tibble(
  gene_id = c(
    # Alpha-mannan backbone cleavage (hemicellulose)
    "GH99", "GH76",
    
    # Alpha-mannan oligo cleavage (hemicellulose)
    "GH125", "GH38", "GH47", "GH63", "GH92", "GH76", "GH130",
    
    # Beta-mannan backbone cleavage (hemicellulose)
    "GH134", "GH113",
    
    # Beta-mannan oligo cleavage (hemicellulose)
    "GH130", "CE2", "GH113",
    
    # Mixed-linkage glucans backbone cleavage (hemicellulose)
    "GH152", "GH157", "GH158", "GH64", "GH81", "GH17", "GH55", "GH128",
    
    # Mixed-linkage glucans oligo cleavage (hemicellulose)
    "GH17", "GH55",
    
    # Xylan backbone cleavage (hemicellulose)
    "GH11", "GH98", "GH26", "GH141", "GH30",
    
    # Xylan oligo cleavage (hemicellulose)
    "GH115", "GH120", "GH39", "GH52", "GH67",
    
    # Xyloglucan backbone cleavage (hemicellulose)
    "GH74",
    
    # Xyloglucan oligo cleavage (hemicellulose)
    "GH31"
    
    
  ),
  class = c(
    rep("Alpha-mannan backbone cleavage (hemicellulose)", 2),
    rep("Alpha-mannan oligo cleavage (hemicellulose)", 7),
    rep("Beta-mannan backbone cleavage (hemicellulose)", 2),
    rep("Beta-mannan oligo cleavage (hemicellulose)", 3),
    rep("Mixed-linkage glucans backbone cleavage (hemicellulose)", 8),
    rep("Mixed-linkage glucans oligo cleavage (hemicellulose)", 2),
    rep("Xylan backbone cleavage (hemicellulose)", 5),
    rep("Xylan oligo cleavage (hemicellulose)", 5),
    rep("Xyloglucan oligo cleavage (hemicellulose)", 1),
    rep("Xyloglucan oligo cleavage (hemicellulose)", 1)
  )
)

```


Classifications (lenient)
```{r}

gene_classes_hem_len <- tibble(
  gene_id = c(
    # Alpha-mannan backbone cleavage (hemicellulose)
    "GH99", "GH76",
    
    # Alpha-mannan oligo cleavage (hemicellulose)
    "GH125", "GH38", "GH47", "GH63", "GH92",
    
    # Beta-mannan backbone cleavage (hemicellulose)
    "GH134", "GH113", "GH26", 
    
    # Beta-mannan oligo cleavage (hemicellulose)
    "GH130", "CE2", "GH2", "GH113", "GH26",
    
    # Mixed-linkage glucans backbone cleavage (hemicellulose)
    "GH152", "GH157", "GH158", "GH64", "GH81", "GH17", "GH55", "GH128", "GH8", "GH16", "GH7", "GH30",
    
    # Mixed-linkage glucans oligo cleavage (hemicellulose)
    "GH1", "GH3", "GH16", "GH17", "GH55",
    
    # Xylan backbone cleavage (hemicellulose)
    "GH11", "GH98", "GH26", "GH141", "GH30",
    
    # Xylan oligo cleavage (hemicellulose)
    "GH115", "GH120", "GH39", "GH52", "GH67", "GH3", "GH31",
    
    # Xyloglucan backbone cleavage (hemicellulose)
    "GH74", "GH9", "GH5", "GH12", "GH44", "GH45", "GH16", "AA9", "GH48",
    
    # Xyloglucan oligo cleavage (hemicellulose)
    "GH31", "GH10", "GH2", "GH43"
    
    
  ),
  class = c(
    rep("Alpha-mannan backbone cleavage (hemicellulose)", 2),
    rep("Alpha-mannan oligo cleavage (hemicellulose)", 5),
    rep("Beta-mannan backbone cleavage (hemicellulose)", 3),
    rep("Beta-mannan oligo cleavage (hemicellulose)", 5),
    rep("Mixed-linkage glucans backbone cleavage (hemicellulose)", 12),
    rep("Mixed-linkage glucans oligo cleavage (hemicellulose)", 5),
    rep("Xylan backbone cleavage (hemicellulose)", 5),
    rep("Xylan oligo cleavage (hemicellulose)", 7),
    rep("Xyloglucan backbone cleavage (hemicellulose)", 9),
    rep("Xyloglucan oligo cleavage (hemicellulose)", 4)
  )
)



```


Obtain target gene_ids for target MAGs
```{r}

# Filter for hemicellulose CAZYs (lenient)
gene_count_meso_hem_len <- gene_counts_meso %>%
  filter(gene_id %in% gene_classes_hem_len$gene_id)

gene_count_thermo_hem_len <- gene_counts_thermo %>%
  filter(gene_id %in% gene_classes_hem_len$gene_id)


# Filter for cellulose CAZYs (conservative)
gene_count_meso_hem_con <- gene_counts_meso %>%
  filter(gene_id %in% gene_classes_hem_con$gene_id)

gene_count_thermo_hem_con <- gene_counts_thermo %>%
  filter(gene_id %in% gene_classes_hem_con$gene_id)



```


--------------------------------------------------------------------------------
                                     Heatmap
--------------------------------------------------------------------------------

Note regarding the (lenient) heatmaps: The most prominent CAZy genes, GH2 and GH3 are multi-functional, encoding among others pectin deg and arabinan deg, making their interpretation for the relevance in cellulose deg impossible. They are perhaps not really involved in cellulose deg.


conservative
```{r}



meso_long <- gene_count_meso_hem_con %>%
  pivot_longer(
    cols = -gene_id,
    names_to = "genome",
    values_to = "count"
  )


thermo_long <- gene_count_thermo_hem_con %>%
  pivot_longer(
    cols = -gene_id,
    names_to = "genome",
    values_to = "count"
  )


p_meso <- ggplot(meso_long, aes(x = genome, y = gene_id, fill = count)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Count", limits = c(0, 15)) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10),
    axis.title = element_blank()
  ) +
  ggtitle("Hemicellulose CAZy repertoire – Mesophilic MAGs")


p_thermo <- ggplot(thermo_long, aes(x = genome, y = gene_id, fill = count)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Count", limits = c(0, 8)) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10),
    axis.title = element_blank()
  ) +
  ggtitle("Hemicellulose CAZy repertoire – Thermophilic MAGs")

p <- p_meso / p_thermo

ggsave(filename = "hemicellulose/gene_count_heatmap_conservative.png", plot = p,
       width = 20, height = 12, dpi = 300)

```


Lenient
```{r}



meso_long <- gene_count_meso_hem_len %>%
  pivot_longer(
    cols = -gene_id,
    names_to = "genome",
    values_to = "count"
  )


thermo_long <- gene_count_thermo_hem_len %>%
  pivot_longer(
    cols = -gene_id,
    names_to = "genome",
    values_to = "count"
  )


p_meso <- ggplot(meso_long, aes(x = genome, y = gene_id, fill = count)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Count", limits = c(0, 15)) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10),
    axis.title = element_blank()
  ) +
  ggtitle("Hemicellulose CAZy repertoire – Mesophilic MAGs")


p_thermo <- ggplot(thermo_long, aes(x = genome, y = gene_id, fill = count)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Count", limits = c(0, 8)) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10),
    axis.title = element_blank()
  ) +
  ggtitle("Hemicellulose CAZy repertoire – Thermophilic MAGs")

p <- p_meso / p_thermo

ggsave(filename = "hemicellulose/gene_count_heatmap_lenient.png", plot = p,
       width = 20, height = 12, dpi = 300)

```


--------------------------------------------------------------------------------
                                     Barplots
--------------------------------------------------------------------------------


Class-separated
```{r}
### Lenient


# ------------------------------------------------------------
# Helper: count MAGs with count > 0 per gene_id
# ------------------------------------------------------------
count_mags_per_gene <- function(df_counts) {
  df_counts %>%
    mutate(across(-gene_id, ~ as.numeric(.x))) %>%
    pivot_longer(-gene_id, names_to = "Genome", values_to = "count") %>%
    mutate(present = !is.na(count) & count > 0) %>%
    group_by(gene_id) %>%
    summarise(n_MAGs = sum(present), .groups = "drop")
}

# helper: make labels fit in skinny facets
stack_gene_label <- function(x) {
  x <- as.character(x)
  gsub("^([A-Z]+)([0-9]+)$", "\\1\n\\2", x)
}

# ------------------------------------------------------------
# Plot function: facet by class + class colours from a gradient
# ------------------------------------------------------------
plot_cazy_by_class <- function(df_counts, gene_classes_hem_len, title, y_max = NULL,
                               gradient_anchors = c("grey80", "steelblue4"),
                               label_stack = TRUE) {

  hits <- count_mags_per_gene(df_counts)

  gene_order <- gene_classes_hem_len %>%
    mutate(gene_id = as.character(gene_id)) %>%
    pull(gene_id)

  plot_df <- gene_classes_hem_len %>%
    mutate(
      gene_id = as.character(gene_id),
      class   = str_squish(class)
    ) %>%
    left_join(hits, by = "gene_id") %>%
    mutate(
      n_MAGs  = replace_na(n_MAGs, 0L),
      class   = fct_inorder(class),
      gene_id = factor(gene_id, levels = unique(gene_order))
    )

  class_levels <- gene_classes_hem_len %>%
    mutate(class = str_squish(class)) %>%
    distinct(class) %>%
    pull(class)

  plot_df <- plot_df %>%
    mutate(class = factor(as.character(class), levels = class_levels))

  pal_fun <- colorRampPalette(gradient_anchors)
  class_cols <- setNames(pal_fun(length(class_levels)), class_levels)

  p <- ggplot(plot_df, aes(x = gene_id, y = n_MAGs, fill = class)) +
    geom_col(width = 0.9) +
    facet_grid(. ~ class, scales = "free_x", space = "free_x", switch = "x") +
    scale_fill_manual(values = class_cols, drop = FALSE) +
    geom_text(aes(label = n_MAGs), vjust = -0.3, size = 2.5) +
    labs(
      x = NULL,
      y = "Number of MAGs (gene count > 0)",
      title = title,
      fill = "Class"
    ) +
    theme_bw() +
    theme(
      axis.text.x = element_text(
        angle = 0,
        hjust = 0.5, vjust = 1,
        size = 7,
        lineheight = 0.9
      ),
      axis.ticks.x = element_line(linewidth = 0.25),
      panel.spacing.x = unit(0.9, "lines"),
      legend.position = "none",
      strip.placement = "outside",
      strip.background = element_rect(fill = "grey95", colour = NA),
      plot.margin = margin(t = 5.5, r = 5.5, b = 18, l = 5.5)
    ) +
    scale_y_continuous(
      limit = c(0, 50),
      expand = expansion(mult = c(0, 0.10))
    ) +
    coord_cartesian(clip = "off")

  if (isTRUE(label_stack)) {
    p <- p + scale_x_discrete(labels = stack_gene_label)
  } else {
    p <- p + scale_x_discrete(drop = FALSE)
  }

  p
}

# ------------------------------------------------------------
# Run for meso (blue gradient) and thermo (red gradient)
# using gene_classes_cel_len
# ------------------------------------------------------------
p_meso <- plot_cazy_by_class(
  df_counts = gene_count_meso_hem_len,        # <-- your cellulose (len) count table
  gene_classes_hem_len = gene_classes_hem_len,
  title = "Number of mesophilic MAGs encoding each cellulose CAZy (count > 0)",
  y_max = 20,
  gradient_anchors = c("cornflowerblue", "royalblue2", "royalblue3", "royalblue4", "midnightblue"),
  label_stack = TRUE
)

p_thermo <- plot_cazy_by_class(
  df_counts = gene_count_thermo_hem_len,      # <-- your cellulose (len) count table
  gene_classes_hem_len = gene_classes_hem_len,
  title = "Number of thermophilic MAGs encoding each cellulose CAZy (count > 0)",
  y_max = 20,
  gradient_anchors = c("sandybrown", "sienna1", "sienna2", "sienna3", "sienna", "sienna4"),
  label_stack = TRUE
)

p <- p_meso / p_thermo
p



ggsave(filename = "hemicellulose/mag_gene_counts_barplot_lenient.png", plot = p,
       width = 20, height = 12, dpi = 300)





### Conservative

# ------------------------------------------------------------
# Helper: count MAGs with count > 0 per gene_id
# ------------------------------------------------------------
count_mags_per_gene <- function(df_counts) {
  df_counts %>%
    mutate(across(-gene_id, ~ as.numeric(.x))) %>%
    pivot_longer(-gene_id, names_to = "Genome", values_to = "count") %>%
    mutate(present = !is.na(count) & count > 0) %>%
    group_by(gene_id) %>%
    summarise(n_MAGs = sum(present), .groups = "drop")
}

# helper: make labels fit in skinny facets
stack_gene_label <- function(x) {
  x <- as.character(x)
  gsub("^([A-Z]+)([0-9]+)$", "\\1\n\\2", x)
}

# ------------------------------------------------------------
# Plot function: facet by class + class colours from a gradient
# ------------------------------------------------------------
plot_cazy_by_class <- function(df_counts, gene_classes_hem_con, title, y_max = NULL,
                               gradient_anchors = c("grey80", "steelblue4"),
                               label_stack = TRUE) {

  hits <- count_mags_per_gene(df_counts)

  gene_order <- gene_classes_hem_con %>%
    mutate(gene_id = as.character(gene_id)) %>%
    pull(gene_id)

  plot_df <- gene_classes_hem_con %>%
    mutate(
      gene_id = as.character(gene_id),
      class   = str_squish(class)
    ) %>%
    left_join(hits, by = "gene_id") %>%
    mutate(
      n_MAGs  = replace_na(n_MAGs, 0L),
      class   = fct_inorder(class),
      gene_id = factor(gene_id, levels = unique(gene_order))
    )

  class_levels <- gene_classes_hem_con %>%
    mutate(class = str_squish(class)) %>%
    distinct(class) %>%
    pull(class)

  plot_df <- plot_df %>%
    mutate(class = factor(as.character(class), levels = class_levels))

  pal_fun <- colorRampPalette(gradient_anchors)
  class_cols <- setNames(pal_fun(length(class_levels)), class_levels)

  p <- ggplot(plot_df, aes(x = gene_id, y = n_MAGs, fill = class)) +
    geom_col(width = 0.9) +
    facet_grid(. ~ class, scales = "free_x", space = "free_x", switch = "x") +
    scale_fill_manual(values = class_cols, drop = FALSE) +
    geom_text(aes(label = n_MAGs), vjust = -0.3, size = 2.5) +
    labs(
      x = NULL,
      y = "Number of MAGs (gene count > 0)",
      title = title,
      fill = "Class"
    ) +
    theme_bw() +
    theme(
      axis.text.x = element_text(
        angle = 0,
        hjust = 0.5, vjust = 1,
        size = 7,
        lineheight = 0.9
      ),
      axis.ticks.x = element_line(linewidth = 0.25),
      panel.spacing.x = unit(0.9, "lines"),
      legend.position = "none",
      strip.placement = "outside",
      strip.background = element_rect(fill = "grey95", colour = NA),
      plot.margin = margin(t = 5.5, r = 5.5, b = 18, l = 5.5)
    ) +
    scale_y_continuous(
      limit = c(0, 25),
      expand = expansion(mult = c(0, 0.10))
    ) +
    coord_cartesian(clip = "off")

  if (isTRUE(label_stack)) {
    p <- p + scale_x_discrete(labels = stack_gene_label)
  } else {
    p <- p + scale_x_discrete(drop = FALSE)
  }

  p
}

# ------------------------------------------------------------
# Run for meso (blue gradient) and thermo (red gradient)
# using gene_classes_cel_len
# ------------------------------------------------------------
p_meso <- plot_cazy_by_class(
  df_counts = gene_count_meso_hem_con,
  gene_classes_hem_con = gene_classes_hem_con,
  title = "Number of mesophilic MAGs encoding each cellulose CAZy (count > 0) (conservative)",
  y_max = 20,
  gradient_anchors = c("cornflowerblue", "royalblue2", "royalblue3", "royalblue4", "midnightblue"),
  label_stack = TRUE
)

p_thermo <- plot_cazy_by_class(
  df_counts = gene_count_thermo_hem_con,      # <-- your cellulose (len) count table
  gene_classes_hem_con = gene_classes_hem_con,
  title = "Number of thermophilic MAGs encoding each cellulose CAZy (count > 0) (conservative)",
  y_max = 20,
  gradient_anchors = c("sandybrown", "sienna1", "sienna2", "sienna3", "sienna", "sienna4"),
  label_stack = TRUE
)

p <- p_meso / p_thermo
p



ggsave(filename = "hemicellulose/mag_gene_counts_barplot_conservative.png", plot = p,
       width = 20, height = 12, dpi = 300)



```


Attribution plot
```{r}


make_stacked_contrib_plot <- function(df_counts,
                                      gene_classes,
                                      top_n = 10,
                                      title = "",
                                      mode = c("proportion", "count"),
                                      facet_by_class = TRUE,
                                      genome_palette = NULL,
                                      other_colour = "grey70",
                                      y_lim = NULL,
                                      mags_tax = NULL,
                                      name_col = "new",
                                      rel_abun_col = "rel_abun",
                                      rel_abun_digits = 2,
                                      select_top_by = c("gene_count", "abundance")) {

  mode <- match.arg(mode)
  select_top_by <- match.arg(select_top_by)

  long <- df_counts %>%
    mutate(across(-gene_id, ~ as.numeric(.x))) %>%
    pivot_longer(-gene_id, names_to = "Genome", values_to = "count") %>%
    mutate(count = replace_na(count, 0))

  present_genomes <- unique(long$Genome)

  if (select_top_by == "abundance") {
    if (is.null(mags_tax)) stop("select_top_by='abundance' requires mags_tax.")

    ab_map <- mags_tax %>%
      select(Genome = all_of(name_col), rel_abun = all_of(rel_abun_col)) %>%
      filter(!is.na(Genome)) %>%
      mutate(rel_abun = as.numeric(rel_abun)) %>%
      distinct(Genome, .keep_all = TRUE)

    keep_genomes <- ab_map %>%
      filter(Genome %in% present_genomes) %>%
      arrange(desc(rel_abun)) %>%
      slice_head(n = top_n) %>%
      pull(Genome)

  } else {
    keep_genomes <- long %>%
      group_by(Genome) %>%
      summarise(total = sum(count), .groups = "drop") %>%
      arrange(desc(total)) %>%
      slice_head(n = top_n) %>%
      pull(Genome)
  }

  long2 <- long %>%
    mutate(Genome_grp = if_else(Genome %in% keep_genomes, Genome, "Other"))

  plot_df <- long2 %>%
    group_by(gene_id, Genome_grp) %>%
    summarise(count = sum(count), .groups = "drop")

  if (mode == "proportion") {
    plot_df <- plot_df %>%
      group_by(gene_id) %>%
      mutate(denom = sum(count), prop = if_else(denom > 0, count / denom, 0)) %>%
      ungroup() %>%
      select(-denom)
  }

  gene_classes2 <- gene_classes %>%
    mutate(gene_id = as.character(gene_id),
           class = str_squish(class))

  gene_order <- gene_classes2 %>% pull(gene_id)
  class_levels <- gene_classes2 %>% distinct(class) %>% pull(class)

  plot_df <- plot_df %>%
    mutate(gene_id = as.character(gene_id)) %>%
    left_join(gene_classes2, by = "gene_id", relationship = "many-to-many") %>%
    mutate(
      class   = factor(class, levels = class_levels),
      gene_id = factor(gene_id, levels = unique(gene_order))
    )

  genome_levels <- c(keep_genomes, "Other")
  plot_df <- plot_df %>% mutate(Genome_grp = factor(Genome_grp, levels = genome_levels))

  n_keep <- length(keep_genomes)

  if (is.null(genome_palette)) {
    pal_fun <- grDevices::colorRampPalette(c("steelblue4", "goldenrod2", "firebrick3"))
    keep_cols <- setNames(pal_fun(n_keep), keep_genomes)
  } else {
    if (length(genome_palette) < n_keep) stop("Not enough colours supplied in genome_palette.")
    keep_cols <- setNames(genome_palette[seq_len(n_keep)], keep_genomes)
  }

  mag_cols <- c(keep_cols, Other = other_colour)

  legend_labels <- setNames(genome_levels, genome_levels)

  if (!is.null(mags_tax)) {
    ab_map2 <- mags_tax %>%
      select(Genome = all_of(name_col), rel_abun = all_of(rel_abun_col)) %>%
      filter(!is.na(Genome)) %>%
      mutate(rel_abun = as.numeric(rel_abun)) %>%
      distinct(Genome, .keep_all = TRUE)

    ab_vec <- ab_map2$rel_abun
    names(ab_vec) <- ab_map2$Genome

    keep_labels <- vapply(
      keep_genomes,
      function(g) {
        ra <- ab_vec[[g]]
        if (is.null(ra) || is.na(ra)) sprintf("%s (NA%%)", g)
        else sprintf(paste0("%s (%.", rel_abun_digits, "f%%)"), g, ra)
      },
      character(1)
    )
    names(keep_labels) <- keep_genomes
    legend_labels <- c(keep_labels, Other = "Other")
  }

  yvar <- if (mode == "proportion") "prop" else "count"
  ylab <- if (mode == "proportion") "Proportion of total gene count"
  else "Total gene count (summed across selected MAGs)"

  p <- ggplot(plot_df, aes(x = gene_id, y = .data[[yvar]], fill = Genome_grp)) +
    geom_col(width = 0.9) +
    scale_fill_manual(values = mag_cols, breaks = genome_levels, labels = legend_labels, drop = FALSE) +
    labs(x = "CAZy family (gene_id)", y = ylab, title = title, fill = "MAG / species") +
    theme_bw() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 7),
      legend.position = "right"
    ) +
    scale_y_continuous(
      labels = if (mode == "proportion") percent_format(accuracy = 1) else waiver(),
      expand = expansion(mult = c(0, 0.05))
    )

  if (facet_by_class) {
    p <- p +
      facet_grid(. ~ class, scales = "free_x", space = "free_x", switch = "x") +
      theme(
        panel.spacing.x = unit(0.8, "lines"),
        strip.placement = "outside",
        strip.background = element_rect(fill = "grey95", colour = NA)
      )
  }

  if (!is.null(y_lim)) p <- p + coord_cartesian(ylim = y_lim)
  p
}

GENE_CLASSES <- gene_classes_hem_len

PAL_MESO_10 <- c(
  "orchid1", "purple", "slateblue", "slateblue4", "navy",
  "steelblue3", "mediumaquamarine", "seagreen1", "seagreen3", "olivedrab"
)

PAL_THERMO_10 <- c(
  "black", "firebrick4", "firebrick3",
  "firebrick1", "gold3", "gold1", "darkorange1",
  "chocolate", "darkorange4", "darkred"
)

# top 10 by gene count (lenient)
p_meso_gene_len <- make_stacked_contrib_plot(
  df_counts = gene_count_meso_hem_len,
  gene_classes = GENE_CLASSES,
  top_n = 10,
  select_top_by = "gene_count",
  mode = "count",
  y_lim = c(0, 140),
  mags_tax = meso_mags_tax,
  name_col = "new",
  rel_abun_col = "rel_abun",
  rel_abun_digits = 2,
  genome_palette = PAL_MESO_10,
  other_colour = "grey80",
  title = "Mesophilic contribution (hemicellulose CAZy gene counts) (lenient) — top 10 by gene count"
)

p_thermo_gene_len <- make_stacked_contrib_plot(
  df_counts = gene_count_thermo_hem_len,
  gene_classes = GENE_CLASSES,
  top_n = 10,
  select_top_by = "gene_count",
  mode = "count",
  y_lim = c(0, 140),
  mags_tax = thermo_mags_tax,
  name_col = "new",
  rel_abun_col = "rel_abun",
  rel_abun_digits = 2,
  genome_palette = PAL_THERMO_10,
  other_colour = "grey80",
  title = "Thermophilic contribution (hemicellulose CAZy gene counts) (lenient) — top 10 by gene count"
)

p_gene_len <- p_meso_gene_len / p_thermo_gene_len
ggsave("hemicellulose/attribution_barplot_lenient_top10_gene.png", p_gene_len,
       width = 20, height = 12, dpi = 300)

# top 10 by gene count (conservative)
p_meso_gene_con <- make_stacked_contrib_plot(
  df_counts = gene_count_meso_hem_con,
  gene_classes = GENE_CLASSES,
  top_n = 10,
  select_top_by = "gene_count",
  mode = "count",
  y_lim = c(0, 35),
  mags_tax = meso_mags_tax,
  name_col = "new",
  rel_abun_col = "rel_abun",
  rel_abun_digits = 2,
  genome_palette = PAL_MESO_10,
  other_colour = "grey80",
  title = "Mesophilic contribution (hemicellulose CAZy gene counts) (conservative) — top 10 by gene count"
)

p_thermo_gene_con <- make_stacked_contrib_plot(
  df_counts = gene_count_thermo_hem_con,
  gene_classes = GENE_CLASSES,
  top_n = 10,
  select_top_by = "gene_count",
  mode = "count",
  y_lim = c(0, 35),
  mags_tax = thermo_mags_tax,
  name_col = "new",
  rel_abun_col = "rel_abun",
  rel_abun_digits = 2,
  genome_palette = PAL_THERMO_10,
  other_colour = "grey80",
  title = "Thermophilic contribution (hemicellulose CAZy gene counts) (conservative) — top 10 by gene count"
)

p_gene_con <- p_meso_gene_con / p_thermo_gene_con
ggsave("hemicellulose/attribution_barplot_conservative_top10_gene.png", p_gene_con,
       width = 20, height = 12, dpi = 300)

# top 10 by abundance (lenient)
p_meso_abun_len <- make_stacked_contrib_plot(
  df_counts = gene_count_meso_hem_len,
  gene_classes = GENE_CLASSES,
  top_n = 10,
  select_top_by = "abundance",
  mode = "count",
  y_lim = c(0, 140),
  mags_tax = meso_mags_tax,
  name_col = "new",
  rel_abun_col = "rel_abun",
  rel_abun_digits = 2,
  genome_palette = PAL_MESO_10,
  other_colour = "grey80",
  title = "Mesophilic contribution (hemicellulose CAZy gene counts) (lenient) — top 10 by abundance"
)

p_thermo_abun_len <- make_stacked_contrib_plot(
  df_counts = gene_count_thermo_hem_len,
  gene_classes = GENE_CLASSES,
  top_n = 10,
  select_top_by = "abundance",
  mode = "count",
  y_lim = c(0, 140),
  mags_tax = thermo_mags_tax,
  name_col = "new",
  rel_abun_col = "rel_abun",
  rel_abun_digits = 2,
  genome_palette = PAL_THERMO_10,
  other_colour = "grey80",
  title = "Thermophilic contribution (hemicellulose CAZy gene counts) (lenient) — top 10 by abundance"
)

p_abun_len <- p_meso_abun_len / p_thermo_abun_len
ggsave("hemicellulose/abun_attribution_barplot_lenient_top10.png", p_abun_len,
       width = 20, height = 12, dpi = 300)

# top 10 by abundance (conservative)
p_meso_abun_con <- make_stacked_contrib_plot(
  df_counts = gene_count_meso_hem_con,
  gene_classes = GENE_CLASSES,
  top_n = 10,
  select_top_by = "abundance",
  mode = "count",
  y_lim = c(0, 35),
  mags_tax = meso_mags_tax,
  name_col = "new",
  rel_abun_col = "rel_abun",
  rel_abun_digits = 2,
  genome_palette = PAL_MESO_10,
  other_colour = "grey80",
  title = "Mesophilic contribution (hemicellulose CAZy gene counts) (conservative) — top 10 by abundance"
)

p_thermo_abun_con <- make_stacked_contrib_plot(
  df_counts = gene_count_thermo_hem_con,
  gene_classes = GENE_CLASSES,
  top_n = 10,
  select_top_by = "abundance",
  mode = "count",
  y_lim = c(0, 35),
  mags_tax = thermo_mags_tax,
  name_col = "new",
  rel_abun_col = "rel_abun",
  rel_abun_digits = 2,
  genome_palette = PAL_THERMO_10,
  other_colour = "grey80",
  title = "Thermophilic contribution (hemicellulose CAZy gene counts) (conservative) — top 10 by abundance"
)

p_abun_con <- p_meso_abun_con / p_thermo_abun_con
ggsave("hemicellulose/abun_attribution_barplot_conservative_top10.png", p_abun_con,
       width = 20, height = 12, dpi = 300)

```


Masked abundance stacked barplot
```{r}


make_abun_stack_plot <- function(df_counts,
                                 gene_classes,
                                 mags_tax,
                                 name_col = "new",
                                 rel_abun_col = "rel_abun",
                                 rel_abun_digits = 2,
                                 top_n = 10,
                                 title = "",
                                 y_lim = NULL,
                                 other_colour = "grey80") {

  long <- df_counts %>%
    mutate(across(-gene_id, ~ as.numeric(.x))) %>%
    pivot_longer(-gene_id, names_to = "Genome", values_to = "count") %>%
    mutate(count = replace_na(count, 0))

  ab_map <- mags_tax %>%
    select(Genome = all_of(name_col),
           rel_abun = all_of(rel_abun_col)) %>%
    filter(!is.na(Genome)) %>%
    mutate(rel_abun = as.numeric(rel_abun)) %>%
    distinct(Genome, .keep_all = TRUE)

  long <- long %>% left_join(ab_map, by = "Genome")

  plot_df0 <- long %>%
    filter(count >= 1) %>%
    select(gene_id, Genome, rel_abun)

  genomes_with_any_gene <- plot_df0 %>%
    distinct(Genome) %>%
    pull(Genome)

  top_genomes <- ab_map %>%
    filter(Genome %in% genomes_with_any_gene) %>%
    arrange(desc(rel_abun)) %>%
    slice_head(n = top_n) %>%
    pull(Genome)

  plot_df <- plot_df0 %>%
    mutate(Genome_grp = if_else(Genome %in% top_genomes, Genome, "Other")) %>%
    group_by(gene_id, Genome_grp) %>%
    summarise(rel_abun = sum(rel_abun, na.rm = TRUE), .groups = "drop")

  gene_classes2 <- gene_classes %>%
    mutate(gene_id = as.character(gene_id),
           class = str_squish(class))

  gene_order <- gene_classes2 %>% pull(gene_id)
  class_levels <- gene_classes2 %>% distinct(class) %>% pull(class)

  plot_df <- plot_df %>%
    mutate(gene_id = as.character(gene_id)) %>%
    left_join(gene_classes2, by = "gene_id", relationship = "many-to-many") %>%
    mutate(
      class   = factor(class, levels = class_levels),
      gene_id = factor(gene_id, levels = unique(gene_order))
    )

  genome_levels <- c(top_genomes, "Other")
  plot_df <- plot_df %>% mutate(Genome_grp = factor(Genome_grp, levels = genome_levels))

  keep_cols <- setNames(scico(length(top_genomes), palette = "managua"), top_genomes)
  mag_cols <- c(keep_cols, Other = other_colour)

  ab_vec <- ab_map$rel_abun
  names(ab_vec) <- ab_map$Genome

  keep_labels <- vapply(
    top_genomes,
    function(g) {
      ra <- ab_vec[[g]]
      if (is.null(ra) || is.na(ra)) sprintf("%s (NA%%)", g)
      else sprintf(paste0("%s (%.", rel_abun_digits, "f%%)"), g, ra)
    },
    character(1)
  )
  names(keep_labels) <- top_genomes
  legend_labels <- c(keep_labels, Other = "Other")

  p <- ggplot(plot_df, aes(x = gene_id, y = rel_abun, fill = Genome_grp)) +
    geom_col(width = 0.9) +
    scale_fill_manual(
      values = mag_cols,
      breaks = genome_levels,
      labels = legend_labels,
      drop = FALSE
    ) +
    labs(
      x = "CAZy family (gene_id)",
      y = "Relative abundance (%)",
      title = title,
      fill = "MAG / species"
    ) +
    theme_bw() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 7),
      legend.position = "right"
    ) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.05)))

  p <- p +
    facet_grid(. ~ class, scales = "free_x", space = "free_x", switch = "x") +
    theme(
      panel.spacing.x = unit(0.8, "lines"),
      strip.placement = "outside",
      strip.background = element_rect(fill = "grey95", colour = NA)
    )

  if (!is.null(y_lim)) p <- p + coord_cartesian(ylim = y_lim)
  p
}

GENE_CLASSES <- gene_classes_hem_len

p_meso_len <- make_abun_stack_plot(
  df_counts = gene_count_meso_hem_len,
  gene_classes = GENE_CLASSES,
  mags_tax = meso_mags_tax,
  name_col = "new",
  rel_abun_col = "rel_abun",
  rel_abun_digits = 2,
  top_n = 10,
  y_lim = c(0, 24),
  title = "Mesophilic (lenient): relative abundance of genomes with CAZy count ≥ 1"
)

p_thermo_len <- make_abun_stack_plot(
  df_counts = gene_count_thermo_hem_len,
  gene_classes = GENE_CLASSES,
  mags_tax = thermo_mags_tax,
  name_col = "new",
  rel_abun_col = "rel_abun",
  rel_abun_digits = 2,
  top_n = 10,
  y_lim = c(0, 24),
  title = "Thermophilic (lenient): relative abundance of genomes with CAZy count ≥ 1"
)

p_len <- p_meso_len / p_thermo_len

ggsave("hemicellulose/relabun_attribution_lenient_top10.png",
       p_len, width = 20, height = 12, dpi = 300)

p_meso_con <- make_abun_stack_plot(
  df_counts = gene_count_meso_hem_con,
  gene_classes = GENE_CLASSES,
  mags_tax = meso_mags_tax,
  name_col = "new",
  rel_abun_col = "rel_abun",
  rel_abun_digits = 2,
  top_n = 10,
  y_lim = c(0, 17),
  title = "Mesophilic (conservative): relative abundance of genomes with CAZy count ≥ 1"
)

p_thermo_con <- make_abun_stack_plot(
  df_counts = gene_count_thermo_hem_con,
  gene_classes = GENE_CLASSES,
  mags_tax = thermo_mags_tax,
  name_col = "new",
  rel_abun_col = "rel_abun",
  rel_abun_digits = 2,
  top_n = 10,
  y_lim = c(0, 17),
  title = "Thermophilic (conservative): relative abundance of genomes with CAZy count ≥ 1"
)

p_con <- p_meso_con / p_thermo_con

ggsave("hemicellulose/relabun_attribution_conservative_top10.png",
       p_con, width = 20, height = 12, dpi = 300)



```


```{r}


```

```{r}


```

















