---
title: "taxonomy"
author: "Lasse Steffensen"
date: "2025-12-17"
output: html_document
---

```{r setup, include=FALSE}

# wd
setwd("C:/Users/lasse/Desktop/wd/25/taxonomy")


# Libraries
library(tidyverse)
library(colorRamp2)
library(forcats)
library(stringr)


# Data frames
BC13_bins <- read.csv("C:/Users/lasse/Desktop/wd/25/df/BC13_bins.tsv", sep="\t")
BC14_bins <- read.csv("C:/Users/lasse/Desktop/wd/25/df/BC14_bins.tsv", sep="\t")
ar53.summary_BC13 <- read.csv("C:/Users/lasse/Desktop/wd/25/df/gtdbtk.ar53.summary_BC13.tsv", sep="\t") %>%
  select(user_genome, classification) %>%
  mutate(sample = "Meso")
ar53.summary_BC14 <- read.csv("C:/Users/lasse/Desktop/wd/25/df/gtdbtk.ar53.summary_BC14.tsv", sep="\t") %>%
  select(user_genome, classification) %>%
  mutate(sample = "Thermo")
bac120.summary_BC13 <- read.csv("C:/Users/lasse/Desktop/wd/25/df/gtdbtk.bac120.summary_BC13.tsv", sep="\t") %>%
  select(user_genome, classification) %>%
  mutate(sample = "Meso")
bac120.summary_BC14 <- read.csv("C:/Users/lasse/Desktop/wd/25/df/gtdbtk.bac120.summary_BC14.tsv", sep="\t") %>%
  select(user_genome, classification) %>%
  mutate(sample = "Thermo")

BC13_summary <- bind_rows(ar53.summary_BC13, bac120.summary_BC13)
BC14_summary <- bind_rows(ar53.summary_BC14, bac120.summary_BC14)
summary <- bind_rows(BC13_summary, BC14_summary)


```



```{r}

# Transform bin data frames

BC13_sel <- BC13_bins %>%
  select(bin, genome_size, cov) %>%
  mutate(sample = "Meso")

BC14_sel <- BC14_bins %>%
  select(bin, genome_size, cov) %>%
  mutate(sample = "Thermo")

bins_combined <- bind_rows(BC13_sel, BC14_sel)


```

```{r}

# Calculate relative abundance for each sample

bins_combined <- bins_combined %>%
  mutate(mapped_bases = cov * genome_size) %>%
  group_by(sample) %>%
  mutate(rel_abun = mapped_bases / sum(mapped_bases, na.rm = TRUE)) %>%
  ungroup()


# Join taxonomy with bins

bins_tax <- bins_combined %>%
  left_join(summary, by = c("sample" = "sample", "bin" = "user_genome"))


# Parse taxonomy ranks

bins_tax2 <- bins_tax %>%
  mutate(classification = replace_na(classification, "d_Unclassified")) %>%
  separate(classification,
           into = c("domain","phylum","class","order","family","genus","species"),
           sep=";", fill="right") %>%
  mutate(
    phylum  = str_remove(phylum,  "^p__"),
    class   = str_remove(class,   "^c__"),
    order   = str_remove(order,   "^o__"),
    family  = str_remove(family,  "^f__"),
    genus   = str_remove(genus,   "^g__"),
    species = str_remove(species, "^s__"),
    phylum  = if_else(is.na(phylum) | phylum == "", "Unclassified", phylum)
  )


# Phyla stacked bar plot data frame

phylum_comp <- bins_tax2 %>%
  group_by(sample, phylum) %>%
  summarise(rel_abun = sum(rel_abun, na.rm = TRUE), .groups = "drop")

phylum_comp_plot <- phylum_comp %>%
  mutate(phylum_plot = if_else(rel_abun < 0.000001, "Other", phylum)) %>%
  group_by(sample, phylum_plot) %>%
  summarise(rel_abun = sum(rel_abun), .groups = "drop")





```

```{r}

# Stacked bar plot figure

ggplot(phylum_comp_plot, aes(x = sample, y = rel_abun, fill = phylum_plot)) +
  geom_col(width = 0.8) +
  labs(x = NULL, y = "Relative abundance (mapped-to-bins)", fill = "Phylum") +
  theme_bw()



```

```{r}

# Phylum ordering
phylum_order <- phylum_comp_plot %>%
  group_by(phylum_plot) %>%
  summarise(total = sum(rel_abun), .groups = "drop") %>%
  arrange(desc(total)) %>%
  pull(phylum_plot)

phylum_comp_plot <- phylum_comp_plot %>%
  mutate(phylum_plot = factor(phylum_plot, levels = phylum_order))

library(RColorBrewer)

n_phyla <- length(levels(phylum_comp_plot$phylum_plot))

# Brewer max is 12–20 depending on palette; Set3 is good for taxonomy
base_cols <- brewer.pal(min(n_phyla, 12), "Set3")

# Expand if needed
cols <- colorRampPalette(base_cols)(n_phyla)

names(cols) <- levels(phylum_comp_plot$phylum_plot)

# Force neutral colours
if ("Other" %in% names(cols)) {
  cols["Other"] <- "grey80"
}
if ("Unclassified" %in% names(cols)) {
  cols["Unclassified"] <- "grey60"
}

ggplot(phylum_comp_plot,
       aes(x = sample, y = rel_abun, fill = phylum_plot)) +
  geom_col(width = 0.8, colour = "black", linewidth = 0.2) +
  scale_fill_manual(values = cols) +
  labs(
    x = NULL,
    y = "Relative abundance",
    fill = "Phylum (no rel. abun. threshold)"
  ) +
  theme_bw() +
  theme(
    legend.key.height = unit(0.4, "cm"),
    legend.text = element_text(size = 9),
    legend.title = element_text(size = 10)
  )


```

Missing: 
Thermotogae


Present:
Fibrobacterota
Pseudomonadota (Proteobacteria)
Bacteroidota (Bacteroidetes)
Bacillota (Firmicutes)


Renamed:
Proteobacteria -> Pseudomonadota
Bacteroidetes -> Bacteroidota
Firmicutes -> Bacillota


Not found in literature:
Atribacterota
Halobacteriota
Patescibacteriota



```{r}

# Resolved/unresolved genomes stacked bar chart

ranks <- c(
  domain  = "d__",
  phylum  = "p__",
  class   = "c__",
  order   = "o__",
  family  = "f__",
  genus   = "g__",
  species = "s__"
)

level_status <- imap_dfr(ranks, \(prefix, rank_name) {
  summary %>%
    mutate(
      rank = rank_name,
      token = str_extract(classification, paste0(prefix, "[^;]*")),
      status = case_when(
        is.na(token) ~ "Unresolved",
        token == prefix ~ "Unresolved",
        token == paste0(prefix, "") ~ "Unresolved",
        TRUE ~ "Resolved"
      )
    ) %>%
    count(sample, rank, status)
}) %>%
  group_by(sample, rank) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  mutate(
    status = factor(status, levels = c("Resolved", "Unresolved")),
    rank = factor(rank, levels = names(ranks))
  )

ggplot(level_status, aes(x = sample, y = prop, fill = status)) +
  geom_col(width = 0.75) +
  facet_wrap(~ rank, nrow = 1) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = c(Resolved = "lightblue", Unresolved = "darkblue")) +
  labs(x = NULL, y = "Fraction of genomes", fill = "Assignment") +
  theme_bw() +
  theme(strip.text = element_text(size = 9))


```

```{r}


#  Ensure mapped bases + per-sample relative abundance
bins_combined <- bins_combined %>%
  mutate(
    mapped_bases = cov * genome_size
  ) %>%
  group_by(sample) %>%
  mutate(
    rel_abun = mapped_bases / sum(mapped_bases, na.rm = TRUE)
  ) %>%
  ungroup()

#  join taxonomy 
bins_tax <- bins_combined %>%
  left_join(summary, by = c("sample" = "sample", "bin" = "user_genome"))

# parse phylum
bins_tax2 <- bins_tax %>%
  mutate(
    classification = replace_na(classification, "d__Unclassified;p__Unclassified")
  ) %>%
  separate(
    classification,
    into = c("domain","phylum","class","order","family","genus","species"),
    sep = ";",
    fill = "right",
    remove = FALSE
  ) %>%
  mutate(
    phylum = str_remove(phylum, "^p__"),
    phylum = if_else(is.na(phylum) | phylum == "" | phylum == "Unclassified", "Unclassified", phylum)
  )

#  Phylum composition per sample 
phylum_comp <- bins_tax2 %>%
  group_by(sample, phylum) %>%
  summarise(rel_abun = sum(rel_abun, na.rm = TRUE), .groups = "drop") %>%
  group_by(sample) %>%
  mutate(rel_abun = rel_abun / sum(rel_abun, na.rm = TRUE)) %>%  # re-normalise for safety
  ungroup()


# Handle low relative abundance phyla as 'other'

phylum_comp_plot <- phylum_comp %>%
  group_by(sample) %>%
  mutate(
    phylum = fct_lump_n(phylum, n = 12, w = rel_abun, other_level = "Other")
  ) %>%
  group_by(sample, phylum) %>%
  summarise(rel_abun = sum(rel_abun), .groups = "drop") %>%
  group_by(sample) %>%
  mutate(rel_abun = rel_abun / sum(rel_abun)) %>%
  ungroup()


phylum_comp_labs <- phylum_comp_plot %>%
  group_by(sample) %>%
  arrange(desc(phylum)) %>%
  mutate(
    frac = rel_abun,
    ypos = cumsum(frac) - 0.5 * frac,
    label = if_else(frac >= 0.05, scales::percent(frac, accuracy = 1), "")
  ) %>%
  ungroup()

ggplot(phylum_comp_labs, aes(x = 2, y = rel_abun, fill = phylum)) +
  geom_col(width = 1, colour = "white", linewidth = 0.2) +
  geom_text(aes(y = ypos, label = label), size = 3) +
  coord_polar(theta = "y") +
  xlim(0.5, 2.5) +
  facet_wrap(~ sample) +
  theme_void() +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "right"
  )


```

```{r}

# ---- Recompute per-sample relative abundance (as you already do) ----
bins_combined <- bins_combined %>%
  mutate(mapped_bases = cov * genome_size) %>%
  group_by(sample) %>%
  mutate(rel_abun = mapped_bases / sum(mapped_bases, na.rm = TRUE)) %>%
  ungroup()

# ---- Join taxonomy ----
bins_tax <- bins_combined %>%
  left_join(summary, by = c("sample" = "sample", "bin" = "user_genome"))

# ---- Parse phylum ----
bins_tax2 <- bins_tax %>%
  mutate(classification = replace_na(classification, "d__Unclassified;p__Unclassified")) %>%
  separate(
    classification,
    into = c("domain","phylum","class","order","family","genus","species"),
    sep = ";",
    fill = "right",
    remove = FALSE
  ) %>%
  mutate(
    phylum = str_remove(phylum, "^p__"),
    phylum = if_else(is.na(phylum) | phylum == "" | phylum == "Unclassified", "Unclassified", phylum)
  )

# ---- Phylum composition per sample ----
phylum_comp <- bins_tax2 %>%
  group_by(sample, phylum) %>%
  summarise(rel_abun = sum(rel_abun, na.rm = TRUE), .groups = "drop") %>%
  group_by(sample) %>%
  mutate(rel_abun = rel_abun / sum(rel_abun, na.rm = TRUE)) %>%
  ungroup()

# ---- (Recommended) Lump small phyla to keep the plot readable ----
# Choose either:
#   A) lump within each sample (may create different "Other" composition per sample)
#   B) lump globally based on mean abundance across samples (keeps categories aligned)
#
# I recommend B for nested doughnuts.

# B) Global lumping by mean abundance across samples
keep_phyla <- phylum_comp %>%
  group_by(phylum) %>%
  summarise(mean_abun = mean(rel_abun), .groups = "drop") %>%
  arrange(desc(mean_abun)) %>%
  slice_head(n = 12) %>%                 # keep top N phyla
  pull(phylum)

phylum_comp_plot <- phylum_comp %>%
  mutate(phylum = if_else(phylum %in% keep_phyla, phylum, "Other")) %>%
  group_by(sample, phylum) %>%
  summarise(rel_abun = sum(rel_abun), .groups = "drop") %>%
  group_by(sample) %>%
  mutate(rel_abun = rel_abun / sum(rel_abun)) %>%
  ungroup()

# ---- Ensure identical phylum factor levels across rings (critical for consistent fills) ----
phylum_levels <- phylum_comp_plot %>%
  group_by(phylum) %>%
  summarise(total = sum(rel_abun), .groups = "drop") %>%
  arrange(desc(total)) %>%
  pull(phylum)

phylum_comp_plot <- phylum_comp_plot %>%
  mutate(
    phylum = factor(phylum, levels = phylum_levels),
    ring  = factor(sample, levels = c("Meso", "Thermo"))  # inner, outer order
  )

phylum_comp_plot <- phylum_comp_plot %>%
  mutate(
    ring = if_else(sample == "Meso", 1, 2)  # numeric radius
  )



ggplot(phylum_comp_plot, aes(x = ring, y = rel_abun, fill = phylum)) +
  geom_col(width = 0.95, colour = "white", linewidth = 0.2) +
  coord_polar(theta = "y") +
  xlim(0.4, 2.6) +
  annotate("text", x = 1, y = 0, label = "Meso", fontface = "bold", vjust = 0.5) +
  annotate("text", x = 2, y = 0, label = "Thermo", fontface = "bold", vjust = 0.5) +
  theme_void() +
  theme(legend.position = "right")



```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
